---
title: "BAMF_2026_GitHub"
author: "Chloe Allison Fouilloux"
date: "2025-12-05"
output: html_document
---
---
title: "Inf_Fib_2026"
author: "Chloe Allison Fouilloux"
date: "2025-10-29"
output: html_document
---

Hello, 

Welcome to our very cool, very awesome **immunity/infection** dataset. Here, we look at 8 lakes every month over the course of a year. We sample the "super"model, the threespine stickleback, looking for the presence and intensity of infection by the trophically transmitted tapeworm, *Schistocephalus solidus*. Paired with looking into infection dynamics in natural populations we also are able to get a pulse of immune dynamics within these same hosts by the extent of fibrosis present in the peritoneal cavity of fish. How cool is that!

This look a lot of time to wrangle and code, so I am very happy for everyone to look through everything and provide feedback and insight. Thank you for your light and support! 

Shine bright like a tapeworm. Hugz, C.

- Tetative Manuscript Title: *The timing of infection dynamics and host immune responses 
in a naturally occurring cestode-vertebrate system*

- Authors: Chloe A. Fouilloux1*, Heather Alexander2, Grace Vaziri2, Patrick McNamara2, Emma Polard2, Amanda Hund3, Edith Reyes Contreras3, Penelope Steffen3, Cate Patterson3, Shira Dubin3, Amy Chen3, Gwen Casey3, John Berini III3, Emma Choi4, Noelle Plantier4, Dina Soklovskaya4,  Daniel I. Bolnick4, Jessica L. Hite1

```{r Load Libraries}

library(tidyverse)
library(tidyr)
library(dplyr)
library(ggplot2)
library(forcats)
library(gghighlight)
library(ggpubr)
library(lubridate)
library(ggridges)
library(mclust)
library(purrr)
library(mgcv)
library(cluster)
library(DHARMa)
library(brms)
library(boot)
library(scam)

library(spdep)
library(sjPlot)

library(tidybayes)
library(bayesplot)

```

Set theme for ggplot

```{r Themes}
#Theme set
theme_set(  theme(
                  strip.background = element_rect(fill = "White"),
                  panel.background = element_rect(fill = "white",
                                                  colour = "black"), 
                  panel.grid.major = element_blank(), 
                  panel.grid.minor = element_blank(), 
                  text = element_text(size = 15)))
```

```{r Loading Data}

#These are little fish analyzed at UConn. These little guys are important for cohort analysis and infection prev!
baby_fish <- read_csv("baby_fish.csv")

#These are fish from BAMF : Dissected by HA and should be used for fibrosis scores
fish_sizes <- read_csv("fish_sizes.csv")

```


TIDYING DATA, please read the notes within!

```{r Tidying Data}
baby_fish <- baby_fish %>%
             mutate(Lake = as.factor(Lake), 
                    Season = as.factor(Season),
                    Month.adj = as.factor(Month.adj), 
                    JOIN_ID = as.factor(JOIN_ID))

fish_sizes <- fish_sizes %>%
             mutate(Lake = as.factor(Lake), 
                    Season = as.factor(Season),
                    Month = as.factor(Month),
                    Month.adj = as.factor(Month.adj), 
                    JOIN_ID = as.factor(JOIN_ID), 
                    Sex = as.factor(Sex))


#LET'S GET TIME ALL IN ORDER, SHALL WE?
fish_sizes$Month.adj <- forcats::fct_relevel(fish_sizes$Month.adj, 
                                         c( "JUL.23", "AUG", "SEP", "OCT", 
                                           "NOV", "DEC", "JAN", "FEB", 
                                           "MAR", "APR", "MAY", "JUN", "JUL.24"))

fish_sizes$Season <- forcats::fct_relevel(fish_sizes$Season, 
                                         c("Summer","Fall", 
                                           "Winter", "Spring"))


#Remember here that max_worm_weight for baby_fish is just the total worm weight. None of these weights are yet transmissible (even when summed), and they are just used for data visualization purposes. 

ALL_fish <- bind_rows(
  fish_sizes %>%
    mutate(Data = factor("adult_fish", levels = c("baby_fish", "adult_fish"))),
  
  baby_fish %>%
    mutate(Fibrosis = NA_real_,
           Data = factor("baby_fish", levels = c("baby_fish", "adult_fish"))))

ALL_fish <- ALL_fish %>%
            mutate(intensity_lengthcorr = (Total_Worms/Length),
                   intensity_masscorr = (Total_Worms/Mass), 
                  wormbiomass_fishmass = total_worm_wt/Mass)

```


################################### 

# COHORTS

################################### 

1.  In July–Sept 2023, larger modal group = F0, smaller = F1.
2.  After winter (say, Feb–March), the F0s should start disappearing.
3.  By summer 2024, F1s have grown and a new small group (F2s) may
    appear.
4.  Fit a mixture model to each lake × month subset.
5.  In the first summer (July–Sept 2023), rank by mean length to label
    cohorts (F0 = largest group).

#First of all, lakes have vastly different sized fish
```{r Fish Sizes and Sampling Counts}

#How many fish did we sample every month?
sampling_effort <- ALL_fish %>%
                   group_by(Lake, Month.adj) %>%
                   summarize(n = n())


ALL_fish %>%
mutate(Lake = fct_reorder(Lake, Length, .desc = TRUE)) %>%
ggplot(aes(x = Lake, y = Length)) +
    geom_jitter(size = 3, alpha = 0.2, height = 0, width = 0.1)+
    stat_summary(geom = "pointrange",
                 shape = 21,
                 size = 1,
                 colour = "salmon",
                 fill = "white",
                 fun.data = mean_cl_boot) +
    ylab("Length (mm)")

#Fish sizes between lakes significantly differ from one another (which is one reason why we can't use simple size cutoffs)

#Kruskal-Wallis test here is for non-parametric data.
kruskal.test(Length ~ Lake, data = ALL_fish)

```

#MODEL-BASED CLUSTERING APPROACH BASED ON FINITE NORMAL MIXTURE
MODELLING.

Here we are doing model-based clustering based on fish length. **Using
mclust.** + Gaussian Finite Mixture Mdoels (GMM) + Subsetted early fish,
(G = 1:2) for F0/F1s. Mclust assigns based on BIC. + Assuming that fish
length data are drawn from a mixture of gaussian distrubutions. + Aimed
at discovering inherent groups within data without prior knowledge of
labels

Read more here:
<https://luca-scr.r-universe.dev/mclust/doc/manual.html#Mclust>

GMMs = assumes each fish length (xi) comes from one of G normal
components; Mixture model below where: 

\pi_k = mixture proportion (the fraction of fish in cohort K, relative abundance); this is the prior probability that a random observation belongs to cohort K **(before seeing the data!)**

\mu_k = mean length of cohort k

\sigma_k\^{2} = variance within that cohort

$$
p(x) \;=\; \sum_{k=1}^{G} \pi_k \,\mathcal{N}\!\bigl(x \mid \mu_k, \sigma_k^{2}\bigr),
\quad
\sum_{k=1}^{G}\pi_k = 1,\; \pi_k \ge 0.
$$

Mclust estimates all these parameters by maximizing the likelihood of our observed data. Then, for each fish, it computes the posterior probability and assignes the fish to whichever K gives the highest probability.

Here, the numerator represents the joint probability that the data point came from cohort K and has a value of x (PRIOR).
The denominator sums those joint probabilities across all possible cohorts (PRIOR: here subscript 'j' instead of 'k' because it runs over all cohorts in the denominator), turning the ratio into a posterior probability (Bayes' theorem).

$$
P(z=k \mid x)
= \frac{\pi_k \,\mathcal{N}\!\bigl(x \mid \mu_k, \sigma_k^{2}\bigr)}
       {\sum_{j=1}^{G} \pi_j \,\mathcal{N}\!\bigl(x \mid \mu_j, \sigma_j^{2}\bigr)}.
$$

We will also be leveraging freely available data about stickleback growth rates from DEB database: https://www.bio.vu.nl/thb/deb/deblab/add_my_pet/entries_web/Gasterosteus_aculeatus/Gasterosteus_aculeatus_res.html 

[WrigWoot2004] Wright, H. A., Wootton, R. J., and Barber, I. (2004). Interpopulation variation in early growth of threespine sticklebacks (Gasterosteus aculeatus) under laboratory conditions. Can. J. Fish. Aquat. Sci., 61:1832--1838.

Wright :: SL = 0.906 * TL + 0.200, r^2 = 0.996 

```{r heck yeah this looks awesome!}

assign_cohorts_bio <- function(ALL_fish_df,
                                f2_months   = 12:13, ## babies in month 12–13, not before!
                                f2_cap_mm   = 25,    #Babies are not bigger than this at 2 months in age 
                               
                                #**** base F1 caps from DEB curve (mm), Month 1..13 
                                f1_cap_mm   = c(30,34,37,40,42,44,45,46,47,48,49,49,49),
                                # lake-specific bumps in mm for selected months
                                #SRA has super big fishies!
                                lake_cap_bump = list(SRA = list(months = 10:13, bump_mm = 20),
                                                     RSS = list(months = 10:13, bump_mm = 4))) {
#Sanity checks on inputs
  stopifnot(all(c("Lake","Month_num","Length","JOIN_ID") %in% names(ALL_fish_df)))
  stopifnot(length(f1_cap_mm) >= max(ALL_fish_df$Month_num, na.rm = TRUE))

#Quick data clean to make sure that everything is the right format and not INFs or NAs
  df <- ALL_fish_df %>%
          mutate(Month_num = as.integer(Month_num), Length = as.numeric(Length)) %>%
          filter(is.finite(Length), is.finite(Month_num))

# ---------- helper to fit a mixture on a single numeric vector ----------
  # helper: up to G comps, sorted means and class ids
  fit_components <- function(x, Gwant) {
    x <- x[is.finite(x)]    # drop non-finite values
    n <- length(x)
    if (n == 0) return(list(means = numeric(0), class = integer(0)))

# number of components cannot exceed number of unique values
    G <- max(1, min(Gwant, n, length(unique(x))))

# try a Gaussian mixture fit with up to G components
# Mclust selects model and G by BIC within the provided range (1:G), where G = a cohort! (seen as a population) 
#GAUSSIAN MIXTURE MODELS: if that throws an error (too few points)s, return a null instead of breaking my beautiful loop
    
    fit <- tryCatch({ if (G == 1) Mclust(x, G = 1) else Mclust(x, G = 1:G) }, error = function(e) NULL)

   # if mixture fit succeeded and has means
    if (!is.null(fit) && !is.null(fit$parameters$mean)) {
      mu  <- as.numeric(fit$parameters$mean) #mean length of cohort K
      ord <- order(mu)              #Order cohort by mean (small to large)
      mu  <- mu[ord]                #sorted means from small to large
  
  # fit$classification are component labels in Mclust’s internal order
  # we remap them into 1..G according to ascending mean so “1 = smallest”
      cl  <- match(fit$classification, ord) 
      #this is a maximum a posteriori (MAP) label = cohort with highest prob. for each fish
      return(list(means = mu, class = cl))
    }

#FOR WINTER MONTHS, DATA DEFICIENT LAKES: Simpler clustering approach, kmeans. Does not model variance or mixture weights
# fallback: kmeans if mixture fails; simpler geometric clustering
    km  <- kmeans(x, centers = G, nstart = 20)
    ord <- order(as.numeric(km$centers))
    mu  <- sort(as.numeric(km$centers))
    cl  <- match(km$cluster, ord)
    list(means = mu, class = cl)
  }
  
    # ---------- end helper ----------

    # ---------- 1) per lake × month: fit and label by rank ----------

  # 1) per lake x month fit + provisional label by rank
  by_month <- df %>%
    group_by(Lake, Month_num) %>%
    group_modify(function(d, key) { # called once per lake-month group
     
       # allow 3 comps only in late months where F2 may appear
      Gwant <- if (key$Month_num %in% f2_months) 3 else 2
      
      fm <- fit_components(d$Length, Gwant) # mixture fit on that month’s lengths
      
      # if no fit (empty group), return placeholders
      if (length(fm$class) == 0)
        return(d %>% mutate(Cohort0 = NA_character_, comp_mean = NA_real_))

      d$comp      <- fm$class
      mu_sorted   <- fm$means
      d$comp_mean <- mu_sorted[d$comp]

      # map components to cohorts based on size rank
      # months 1–11: small → F1, large → F0
      # months 12–13: small → F2, middle → F1, large → F0
      if (length(mu_sorted) == 1) {
        lab_map <- if (key$Month_num <= 3) c("F0") else c("F1")
      } else if (length(mu_sorted) == 2) {
        lab_map <- c("F1","F0")
      } else {
        lab_map <- c("F2","F1","F0")
      }
      d$Cohort0 <- lab_map[d$comp]
      d
    }) %>%
    ungroup()
  
   # ---------- 2) compute monthly cohort targets (means) ----------
  # for each lake-month-cohort, use the mean Length as the target

  # 2) month targets from means
  cohort_means <- by_month %>%
    filter(!is.na(Cohort0)) %>%
    group_by(Lake, Month_num, Cohort0) %>%
    summarise(mu = mean(Length), .groups = "drop") %>%

    # drop any F2 outside late months (safety)
    mutate(Cohort0 = ifelse(Cohort0 == "F2" & !(Month_num %in% f2_months), NA, Cohort0)) %>%
    filter(!is.na(Cohort0))

  # pivot to wide so we have columns F0, F1, F2 with their monthly mean targets
  targets <- cohort_means %>%
    transmute(Lake, Month_num, Cohort = Cohort0, mu_mon = mu) %>%
    tidyr::pivot_wider(names_from = Cohort, values_from = mu_mon)

    # ---------- 3) assign each fish to the nearest target ----------

  # 3) assign to nearest target
  assigned <- df %>%
    left_join(targets, by = c("Lake","Month_num")) %>%
    mutate(
      dF0 = ifelse(is.finite(F0), abs(Length - F0), Inf),
      dF1 = ifelse(is.finite(F1), abs(Length - F1), Inf),
      dF2 = ifelse(is.finite(F2), abs(Length - F2), Inf),
  #this looks at each row, and compares real fish length to the mean of each cluster. 
  #whichever cluster it is closest to, that's the one it will be assigned.
      Cohort = c("F0","F1","F2")[apply(cbind(dF0, dF1, dF2), 1, which.min)]
    )

    # ---------- 4) enforce biological caps (per month, with lake bumps) ----------

  # 4) apply biological caps with optional lake bumps
  # build a per-row cap for F1
  f1_cap_vec <- f1_cap_mm[pmin(assigned$Month_num, length(f1_cap_mm))]

  # apply lake-specific bumps
    # apply lake-specific bumps: for listed lakes and months, add bump_mm to the cap

  if (length(lake_cap_bump)) {
    for (lk in names(lake_cap_bump)) {
      months <- lake_cap_bump[[lk]]$months
      bump   <- lake_cap_bump[[lk]]$bump_mm
      idx <- which(assigned$Lake == lk & assigned$Month_num %in% months)
      if (length(idx)) f1_cap_vec[idx] <- f1_cap_vec[idx] + bump
    }
  }

  assigned <- assigned %>%
    mutate(
      # F2 only in late months and very small
      Cohort = ifelse(Month_num %in% f2_months & Length <= f2_cap_mm, "F2", Cohort),
      Cohort = ifelse(!(Month_num %in% f2_months) & Cohort == "F2", "F1", Cohort),

      # F1 cannot exceed cap -> becomes F0
      Cohort = ifelse(Cohort == "F1" & Length > f1_cap_vec, "F0", Cohort),

      # F0 cannot be below the F1 cap -> becomes F1
      Cohort = ifelse(Cohort == "F0" & Length <= f1_cap_vec, "F1", Cohort),

      Cohort = factor(Cohort, levels = c("F0","F1","F2"))
    ) %>%
    select(JOIN_ID, Lake, Month_num, Length, Cohort)

  assigned
}


###################################################################
##################################################################
ALLfish_with_cohorts <- assign_cohorts_bio(ALL_fish, f2_months = 12:13, f2_cap_mm = 25)

ALL_fish <- ALL_fish %>%
            left_join(ALLfish_with_cohorts %>%
                        select(JOIN_ID, Cohort) %>% 
                        distinct(),
                          by = c("JOIN_ID")) %>%
            drop_na(Cohort)

cohort_data_full <- ALL_fish%>%
                    mutate(DEB = (100*Mass/(Length^3)), na.rm = T) %>%
                    mutate(inf_yn = ifelse(Total_Worms>0, 1, 0)) %>%
                    group_by(Lake) %>% #X LAKE 
                    mutate(
                      inf_prev_lake = sum(inf_yn, na.rm=TRUE)/n()) %>%
                    group_by(Lake, Month.adj) %>% #X LAKE X MONTH
                    mutate(
                      inf_prev_lake_month = sum(inf_yn, na.rm=TRUE)/n()) %>%
                    group_by(Lake, Cohort) %>% # XLAKE X COHORT
                    mutate(
                      inf_prev_lake_cohort = sum(inf_yn, na.rm=TRUE)/n()) %>%
                    ungroup() %>%
                    group_by(Lake, Month.adj, Cohort) %>%  #X LAKE X MONTH X COHORT
                    mutate(
                      inf_prev_lake_month_cohort = sum(inf_yn, na.rm=TRUE)/n()) %>%
                    ungroup() %>%
                    mutate(Season = as.factor(Season)) %>%
                    mutate(hi_low_fib = case_when(
                                  Lake %in% c("SUG", "SRA", "L2B") ~ "High",
                                  Lake %in% c("PCH", "FRE", "BLA", "RSS", "BEW")~ "Low",
                                 TRUE ~ NA_character_), 
                            hi_low_fib = as.factor(hi_low_fib)) 


fish_sizes <- fish_sizes %>%
              left_join(
                cohort_data_full %>% select(JOIN_ID, Cohort) %>% distinct(),
                by = c("JOIN_ID")) 

####### Let's look at this ABSOLUTE BEAUTY OF A PLOT! ####
cohort_data_full %>%
mutate(Lake = fct_reorder(Lake, Fibrosis, .desc = TRUE)) %>%
ggplot(aes(x = Month_num, y = Length, color = Cohort)) +
  geom_point(alpha = 0.4, size = 1) +
#The smoothed line is generated using a shape-constrained additive model (SCAM), where the length–month relationship #is fit with a monotone P-spline (bs = "mpi").
  geom_smooth(method = scam::scam, #NOTE! this is from a seperate package!
              formula = y ~ s(x, k = 4, bs = "mpi"),
              se = T, linewidth = 1.2) +
  facet_wrap(~Lake, ncol = 4) +
  scale_x_continuous(
    breaks = unique(cohort_data_full$Month_num),
    labels = unique(cohort_data_full$Month.adj)) +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(x = "", y = "Length (mm)") +
  scale_colour_manual(values = c("F0" = "#FF6347", 
                                "F1" = "#477187", 
                                "F2" = "lightblue"))


```


################################### 

# Generating clusters of high/low fibrosis lakes

################################### 

 Alternative coding options are as follows, **kmenoids** (PAM) or even, 

```{r Hi-Low Lake Clustering (not int.prev.)}

lake_summary <- fish_sizes %>%
                group_by(Lake) %>%
                summarise(
                  fib_median   = median(Fibrosis, na.rm = TRUE),
                  fib_mean     = mean(Fibrosis, na.rm = TRUE),
                  fib_se =  sd(Fibrosis, na.rm = TRUE)/sqrt(sum(!is.na(Fibrosis))),
                  inf_prev     = mean(Total_Worms > 0, na.rm = TRUE),
                  mean_worm_load = mean(Total_Worms, na.rm = TRUE),
                  ints_se =  sd(Total_Worms, na.rm = TRUE)/sqrt(sum(!is.na(Total_Worms))),
                  .groups = "drop" )

lake_summary
# Then feed fib_median or fib_mean into PAM
#this is giving the highest weight to fibrosis
w <- c(fib_median = 0.5, fib_mean = 0.3,  inf_prev = 0.1, mean_worm_load = 0.1)

#Daily is a dissimilarity matrix for mixed-type data
diss <- daisy( lake_summary %>% select(fib_median, fib_mean, inf_prev, mean_worm_load),
               metric = "gower", stand = TRUE, weights = as.numeric(w))
diss

pam_fit <- pam(diss, k = 2)
lake_summary %>% mutate(cluster = factor(pam_fit$clustering))

###
fish_sizes <- fish_sizes %>% mutate(hi_low_fib = case_when(
                                                Lake %in% c("SUG", "SRA", "L2B") ~ "High",
                                                Lake %in% c("PCH", "FRE", "BLA", "RSS","BEW")~ "Low",
                                               TRUE ~ NA_character_), 
                                    hi_low_fib = as.factor(hi_low_fib))

```

################################### 

# VISUALITIONS

################################### 

#FIBROSIS
Notes. Fibrosis scores are only for adults, not babies. 

 (1) Across lakes, fibrosis peaks in summer months and is higher in older stickleback.
 (2) Fibrosis and infection prevalence (on a monthly scale) do not track with one another
 (3) [Low-fibrosis lakes] fish with larger tapeworms have higher fibrosis scores (maybe trying to suppress the tapeworm growth?)
 (4) [Low-fibrosis lakes] have more transmissible sized worms, and these worms more frequently occur in summer months. 

```{r VISUALIZATION: Fibrosis prevalence for adults}

#####################################
######## FIBROSIS PREVALENCE FOR adults (count) ######
#####################################
fish_sizes %>%
    mutate(Lake = fct_reorder(Lake, Fibrosis, .desc = TRUE)) %>%
  ggplot(aes(x = Lake, colour = hi_low_fib,
             y = Fibrosis)) +
  geom_jitter(width = 0.1, height = 0.08, 
              size = 2,
              alpha = 0.3) +
  stat_summary(geom = "pointrange", 
               fill = "white",
               size = 1.1,
               shape = 21, 
               colour = "black",
               fun.data = mean_cl_boot)+ #mean_cl_boot
  labs(x = "", y = "Mean fibrosis score (0-4)") +
    scale_colour_manual(name = "High/Low Fibrosis Lake", 
                      values = c("#009E73", "grey"))+
  theme(axis.title.y  = element_text(size = 20,  angle = 90))


#With a smoother
#FIBROSIS BY COHORT + HI/LOW FIBROSIS
fish_sizes %>%
  filter(!Cohort %in% c(NA ,"F2")) %>%
  ggplot(aes(x = Month.adj, 
             y = Fibrosis, colour = Cohort)) +
  geom_jitter(width = 0.1, height = 0.08, 
              size = 2,
              alpha = 0.05) +
  stat_smooth(method = "gam", 
              aes(group = Cohort),
              formula = y ~ s(x, bs = "cc", k = 4),
              fullrange = T ) +
  labs(x = "", y = "Fibrosis Score (0-4)") +
    scale_colour_manual(values = c("F0" = "#FF6347", 
                                "F1" = "#477187", 
                                "F2" = "lightblue"))+
  scale_fill_manual(values = c("F0" = "#FF6347", 
                                "F1" = "#477187", 
                                "F2" = "lightblue")) +
  facet_wrap(~as.factor(hi_low_fib),labeller = as_labeller(c(
                                              "High" = "High fibrosis population",
                                              "Low"  = "Low fibrosis population"))) +
  theme(strip.text = element_text(size = 18),
        axis.text.y = element_text(size = 14),
        axis.title.y = element_text(size = 16, angle =90),
        strip.text.x = element_text(margin = margin(0.25,0,0.25,0, "cm"))) +
  theme(legend.position = "None")

#######################
## LAKE-LEVEL VISUALIZATION
###################
fish_sizes %>%
  group_by(Lake) %>%
  mutate(mean_fib = mean(Fibrosis, na.rm = T))%>%
  ungroup() %>%
  mutate(Lake = fct_reorder(Lake, Fibrosis, .desc = TRUE)) %>%
  ggplot(aes(x = Month.adj, 
             y = Fibrosis, colour = Lake)) +
    geom_hline(aes(yintercept =  mean_fib), colour = "red")+
  geom_jitter(width = 0.1, height = 0.08, 
              size = 2, alpha = 0.05) +
  stat_smooth(method = "gam", 
              aes(group = 1),
              formula = y ~ s(x, bs = "cs", k = 6),
              #formula = y ~ poly(x,4),
              fullrange = T )+
  # stat_summary(geom = "line", 
               # aes(group = 1),
               #size = 1,
              # fun.data = mean_cl_boot)+
  facet_wrap(~Lake, ncol = 4)+
  labs(x = "", y = "Fibrosis score (0-4)") +
  theme(axis.title.y  = element_text(size = 20,  angle = 90), 
        axis.text.x =element_text( angle = 90), 
        legend.position = "none") +
  ylim(0, 4)

## LAKE-LEVEL VISUALIZATION

mean_fibrosis <- fish_sizes %>%
  filter(!Cohort == "NA") %>%
  group_by(Lake, Month.adj) %>%
  summarise(mean_fib = mean(Fibrosis, na.rm = TRUE), .groups = "drop")

# For each lake, find the max mean fibrosis (taking into account that BEW has so few observations)
max_points <- mean_fibrosis %>%
  group_by(Lake) %>%
  filter(mean_fib == max(mean_fib)) %>%
  slice_min(Month.adj, n = 1) %>%
  ungroup()

fish_sizes %>%
  filter(!Cohort == "NA") %>%
  group_by(Lake) %>%
  mutate(mean_fib = mean(Fibrosis, na.rm = TRUE)) %>%
  ungroup()%>%
  mutate(Lake = fct_reorder(Lake, mean_fib, .desc = TRUE)) %>%
  ggplot(aes(x = Month.adj, y = Fibrosis)) +
    geom_hline(aes(yintercept =  mean_fib), colour = "red")+
  #highlighting summer months!
    annotate(geom = "rect", 
           xmin = -Inf, xmax = 2.5, 
           ymin = -Inf, ymax = Inf, 
           fill = "#FF8C2C", 
           alpha = 0.3)+
    annotate(geom = "rect", 
           xmin = 11.5, xmax = Inf, 
           ymin = -Inf, ymax = Inf, 
           fill = "#FF8C2C", 
           alpha = 0.3)+
  geom_jitter(width = 0.1, height = 0.08, 
              size = 3, alpha = 0.05) +
  stat_summary(geom = "line", aes(group = 1), 
               size = 1, fun.data = mean_cl_boot) +
  geom_point(data = max_points, 
             aes(x = Month.adj, y = mean_fib), color = "red", size = 3) +
  facet_wrap(~Lake, ncol = 4) +
  labs(x = "", y = "Fibrosis score (0-4)") +
  theme(axis.text.x = element_text(angle = 90))

```

```{r VISUALIZATION: Infection prevalenece for all}
#ALL OF THE FISH, INFECTION PREVALENCE BY COHORT.
cohort_data_full %>%
  filter(!Cohort == "NA") %>%
  ggplot( aes(x = Month.adj,y = inf_prev_lake_month_cohort,
              colour = Cohort, group =Cohort)) +
  geom_jitter(
              width = 0.1, height = 0.08, 
              size = 3, alpha = 0.05) +
  #stat_summary(geom = "line", 
              # size = 1, fun.data = mean_cl_boot) +
    stat_smooth(method = "gam", 
              aes(group = Cohort),
              formula = y ~ s(x, bs = "cs", k = 4),
              #formula = y ~ poly(x,4),
              fullrange = T )+
    scale_colour_manual(values = c("F0" = "#FF6347", 
                                "F1" = "#477187", 
                                "F2" = "lightblue"))+
  scale_fill_manual(values = c("F0" = "#FF6347", 
                                "F1" = "#477187", 
                                "F2" = "lightblue"))+
 ylim(-0.05,1.05)+
 labs(x = "", y = "Infection prevalence by lake by month by cohort") 

###############
# Lake level
##############

#Infection prevalence by lakes, by month. 
cohort_data_full %>%
    mutate(Lake = fct_reorder(Lake, Fibrosis, .desc = TRUE)) %>%
  ggplot(aes(x = Month.adj, y = inf_prev_lake_month, colour = Lake)) +
  geom_jitter(
              width = 0.1, height = 0.08,
              size = 3, alpha = 0.05) +
      stat_smooth(method = "gam", 
              aes(group = 1),
              formula = y ~ s(x, bs = "cs", k = 3),
              #formula = y ~ poly(x,4),
              fullrange = T )+
  facet_wrap(~Lake, ncol = 4)+
  ylim(-0.05,1.05)+
  labs(x = "", y = "Infection prevalence by lake by month") +
  theme(axis.text.x = element_text(angle = 90), 
        legend.position = "none")

```

```{r VISUALIZATIONS Fibrosis and Infection Prevalence | P:H Biomass}

#################################
##################################

#Lake-level analysis
####### WITH SMOOTHING
# Dual-axis plot: Infection prevalence (0–1) and Fibrosis (0–4)
cohort_data_full%>%
  #filter(Lake %in% c("SUG")) %>% #"SUG", "L2B", "SRA"
  mutate(Lake = fct_reorder(Lake, Fibrosis, .desc = TRUE)) %>%
  ggplot(aes(x = Month.adj)) +
  # --- Infection prevalence (left y-axis, 0–1) ---
  geom_jitter(aes(y = inf_yn),
              width = 0.1, height = 0.08,
              colour = "#586E75", size = 2.5, alpha = 0.05) +
  stat_smooth(method = "gam", aes(y = inf_prev_lake_month, group = 1),
              formula = y ~ s(x, bs = "cs", k = 3),
              colour = "black",
              fullrange = T ) +

  # --- Fibrosis (right y-axis, 0–4), rescaled for overlay ---
  # you must plot Fibrosis scaled to 0–1 and then define a secondary axis that multiplies by 4 for the labels
  #-> plot divided by 4 to share 0–1 frame ---
  geom_jitter(aes(y = Fibrosis / 4),
              width = 0.1, height = 0.08,
              colour = "#FFCD04", size = 2.5, alpha = 0.05) +
  stat_smooth(method = "gam", aes(y = Fibrosis/4, group = 1),
              colour = "#D7AC00",
              formula = y ~ s(x, bs = "cs", k = 4),
              fullrange = T ) +

  # --- Facet by fibrosis grouping ---
  facet_wrap( ~ Lake , ncol = 4) +

  # --- Dual y-axes ---
  xlab("")+
  scale_y_continuous(
    name = "Infection prevalence (lake x month)",
    limits = c(0, 1),
    sec.axis = sec_axis(~ . * 4, name = "Fibrosis score (0–4)") ) +
     theme(legend.position = "None", 
        axis.text.y = element_text(size = 15),
        axis.text.x = element_text(size = 15, angle = 90),
        axis.title.x = element_text(size = 22),
        axis.title.y = element_text(size = 22, angle =90),
        axis.title.y.right = element_text(color = "#D7AC00"))

####
#####

# Dual-axis plot: Wormmass:HostMass and Fibrosis (0–4)
fish_sizes %>%
  filter(!Cohort == "F2") %>%
  #filter(wormbiomass_fishmass >0) %>%
  mutate(Lake = fct_reorder(Lake, Fibrosis, .desc = TRUE)) %>%
  ggplot(aes(x = Month.adj)) +

  # --- host parasite biomass (left y-axis, 0–1) ---
  geom_jitter(aes(y = wormbiomass_fishmass),  # max_worm_wt
              width = 0.1, height = 0.08,
              colour = "red", size = 2.5, alpha = 0.05) +
    stat_smooth(method = "gam", aes(y = wormbiomass_fishmass, group = 1), #max_worm_wt
              colour = "salmon",
              fill = "salmon",
              formula = y ~ s(x, bs = "cs", k = 3),
              fullrange = T ) +

  # --- Fibrosis (right y-axis, 0–4), rescaled for overlay ---
  # you must plot Fibrosis scaled to 0–1 and then define a secondary axis that multiplies by 4 for the labels
  #-> plot divided by 4 to share 0–1 frame ---
  geom_jitter(aes(y = Fibrosis / 13),
              width = 0.1, height = 0.08,
              colour = "#FFCD04", size = 2.5, alpha = 0.05) +
    stat_smooth(method = "gam", aes(y = Fibrosis/13, group = 1),
              colour = "#D7AC00",
              fill = "gold",
              formula = y ~ s(x, bs = "cs", k = 3),
              fullrange = T ) +

  # --- Facet by fibrosis grouping ---
  facet_wrap( ~  Lake, ncol = 4) +

  # --- Dual y-axes ---
  scale_y_continuous(
    name = "Parasite:host biomass",
    limits = c(0, 0.3),
    sec.axis = sec_axis(~ . * 13, name = "Fibrosis score (0–4)") ) +
  xlab("")+
  theme(
    axis.title.y.left = element_text(color = "salmon", size = 16),
    axis.title.y.right = element_text(color = "#D7AC00", size = 16),
    axis.text.x = element_text(angle = 90),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank() )

### Just in infected fish
# Dual-axis plot: Wormmass:HostMass and Fibrosis (0–4)
fish_sizes %>%
  filter(!Cohort == "F2") %>%
  mutate(Lake = fct_reorder(Lake, Fibrosis, .desc = TRUE)) %>%
  filter(wormbiomass_fishmass >0) %>%
  ggplot(aes(x = Month.adj)) +

  # --- host parasite biomass (left y-axis, 0–1) ---
  geom_jitter(aes(y = wormbiomass_fishmass),  # max_worm_wt
              width = 0.1, height = 0.08,
              colour = "red", size = 2.5, alpha = 0.05) +
    stat_smooth(method = "gam", aes(y = wormbiomass_fishmass, group = 1), #max_worm_wt
              colour = "salmon",
              fill = "salmon",
              formula = y ~ s(x, bs = "cs", k = 3),
              fullrange = T ) +

  # --- Fibrosis (right y-axis, 0–4), rescaled for overlay ---
  # you must plot Fibrosis scaled to 0–1 and then define a secondary axis that multiplies by 4 for the labels
  #-> plot divided by 4 to share 0–1 frame ---
  geom_jitter(aes(y = Fibrosis / 13),
              width = 0.1, height = 0.08,
              colour = "#FFCD04", size = 2.5, alpha = 0.05) +
    stat_smooth(method = "gam", aes(y = Fibrosis/13, group = 1),
              colour = "#D7AC00",
              fill = "gold",
              formula = y ~ s(x, bs = "cs", k = 3),
              fullrange = T ) +

  # --- Facet by fibrosis grouping ---
  facet_wrap( ~  Lake, ncol = 4) +

  # --- Dual y-axes ---
  scale_y_continuous(
    name = "Parasite:host biomass of infected fish",
    limits = c(0, 0.3),
    sec.axis = sec_axis(~ . * 13, name = "Fibrosis score (0–4)") ) +
  xlab("")+
  theme(
    axis.title.y.left = element_text(color = "salmon", size = 16),
    axis.title.y.right = element_text(color = "#D7AC00", size = 16),
    axis.text.x = element_text(angle = 90),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank() )

  
####

```

```{r MAXIMUM WORM WT TABLES: HI/Low firosis lakes and parasite growth}

table_worm1 <- cohort_data_full %>%
                mutate(hi_low_fib = case_when(
                Lake %in% c("SUG", "SRA", "L2B") ~ "High",
                Lake %in% c("PCH", "FRE", "BLA", "RSS", "BEW") ~ "Low",
                TRUE ~ NA_character_ )) %>%
            mutate(trans_inf = ifelse(max_worm_wt >= 0.05, 1, 0), 
                   hi_low_fib = as.factor(hi_low_fib)) %>%
            group_by(Month.adj, hi_low_fib) %>%
            summarise(n_fish = n(), 
                      n_inf_hilow = sum(trans_inf, na.rm = T),
                      prp_trans_hilow = n_inf_hilow/n_fish) %>%
            ungroup()

print(table_worm1, n = 50)
  
overall_transmissible <- cohort_data_full %>%
  mutate(
    hi_low_fib = case_when(
      Lake %in% c("SUG", "SRA", "L2B") ~ "High",
      Lake %in% c("PCH", "FRE", "BLA", "RSS", "BEW") ~ "Low",
      TRUE ~ NA_character_ ),
    trans_inf = ifelse(max_worm_wt >= 0.05, 1, 0)  ) %>%
  filter(!is.na(hi_low_fib)) %>%      
  group_by(hi_low_fib) %>%
  summarise(
    n_fish = n(),
    n_transmissible = sum(trans_inf, na.rm = TRUE),
    prp_transmissible = n_transmissible / n_fish ) %>%
   ungroup()

overall_transmissible
  
#MAXIMUM WORM WEIGHT with insight on transmissible stages
# Here ya go
threshold <- 0.05  # adjust to match your cutoff line

cohort_data_full %>%
  filter(!Cohort == "F2") %>%
  #filter( !max_worm_wt == 0) %>%
  ggplot(aes(x = Month.adj, 
             y = max_worm_wt, 
             #max_worm_wt = Nice trend
             #log(max_worm_wt+1)= Nice trend
             #(max_worm_wt/Mass)= Nice trend
             #max_worm_wt/Length = Nice trend
             colour = ifelse(max_worm_wt >= 0.05, 
                             Season, "grey"))) + 
  
  geom_jitter(width = 0, height = 0, size = 4,
              alpha = 0.5) +
  #Transmissible worms at 50mg
  geom_hline(yintercept = 0.05)+
   facet_wrap(~ Cohort)+
          # smoother ONLY for points above threshold
  stat_smooth(
            data = ~ dplyr::filter(., max_worm_wt > threshold),
            method = "gam", aes(group = 1),
            formula = y ~ s(x, k = 3),
            se = TRUE, colour = "black", fill = "grey50" ) +
  scale_colour_manual(values = c(
                        "1" = "#FF8C2C",
                        "2" = "#DBBE00",
                        "3" = "#427793",
                        "4" = "#92D057", 
                        "grey" = "lightgrey")) +
   labs(y = "Maximum worm weight (g) in all fish", x = "")+
    theme(
    strip.text  = element_text(size = 22),
    axis.title.y = element_text(size = 20, angle = 90),
    axis.text.y = element_text(size = 20),
    axis.text.x = element_text(size = 12, angle = 90),
    legend.position = "none"  )  

#MAXIMUM WORM WEIGHT with insight on transmissible stages

fib_labels <- c("High" = "High fibrosis lakes",
                "Low" = "Low fibrosis lakes")

cohort_data_full %>%
  mutate(hi_low_fib = case_when(
    Lake %in% c("SUG", "SRA", "L2B") ~ "High",
    Lake %in% c("PCH", "FRE", "BLA", "RSS", "BEW") ~ "Low",
    TRUE ~ NA_character_ )) %>%
 # filter(Total_Worms != 0) %>%
  ggplot(aes(y = max_worm_wt, 
             x = Month.adj, 
             #log(Total_Worms + 1)
             #log(intensity_masscorr+1)
          shape = ifelse(max_worm_wt >= 0.05, 
                            as.factor(hi_low_fib), "Not transmissible"),
           colour = ifelse(max_worm_wt >= 0.05, 
                            as.factor(hi_low_fib), "Not transmissible"))) + 
  geom_point(size = 6, alpha = 0.6) +
  geom_hline(yintercept = 0.05)+
  #stat_smooth(
           # data = ~ dplyr::filter(., max_worm_wt > threshold),
           # method = "gam", aes(group = 1),
           # formula = y ~ s(x, k = 3),
           # se = TRUE, colour = "black", fill = "grey50" ) +
  facet_wrap(~ hi_low_fib, 
             labeller = labeller(hi_low_fib = fib_labels)) +
  labs(x = "", 
       y = "Maximum tapeworm weight in host (grams)") +
  scale_colour_manual(name = "High/Low Fibrosis Lake", 
                        values = c("#009E73", "grey", "lightgrey"))+
  scale_shape_manual(values = c(16, 16, 21))+
 theme(legend.position = "None", 
      strip.text = element_text(size = 18),
        axis.text.y = element_text(size = 18),
        axis.title.y = element_text(size = 20, angle =90),
        strip.text.x = element_text(margin = margin(0.25,0,0.25,0, "cm")))


cohort_data_full %>%
  #filter(Total_Worms != 0) %>%
  ggplot(aes(x = max_worm_wt, colour = hi_low_fib,
             y = Fibrosis)) +
  geom_jitter(size = 5, height = 0.1, width = 0, alpha = 0.4) +
  stat_smooth(method = "gam", 
              fullrange = F,
              formula = y ~ s(x, k = 3, bs = "cs"))+
scale_colour_manual(name = "High/Low Fibrosis Lake", 
                        values = c("#009E73", "grey", "#AECDC870"))+
  labs(y = "Fibrosis score (0-4)", 
       x = "Maximum worm weight in infected hosts (grams)") 

#Infection prevalence is lowest in population with small parasites

plot_df <- 
  cohort_data_full %>%
  group_by(Lake, hi_low_fib) %>%
        mutate(
          mean_wormsize_lake = mean(total_worm_wt, na.rm = TRUE),
          mean_maxworm_lake  = mean(max_worm_wt,    na.rm = TRUE), 
          mean_fib_lake = mean(Fibrosis, na.rm = T)) %>%
        summarise(
            inf_prev_lake      = mean(inf_prev_lake,  na.rm = TRUE),
            mean_maxworm_lake  = mean(mean_maxworm_lake, na.rm = TRUE),
            mean_wormsize_lake = mean(mean_wormsize_lake, na.rm = T),
            mean_fib_lake = mean(mean_fib_lake, na.rm = T),
            se_max             = sd(max_worm_wt, na.rm = TRUE) / sqrt(n()))

ggplot(plot_df, aes(x = inf_prev_lake, y = mean_maxworm_lake, colour = hi_low_fib)) +
        geom_point(size = 3) +
        geom_errorbar(aes(
          ymin = mean_maxworm_lake - se_max,
          ymax = mean_maxworm_lake + se_max ), width = 0) +
      stat_cor(aes(group = 1))+
      stat_smooth(method = "gam", aes(group = 1), colour = "black",
                    formula = y ~ s(x, bs = "cs", k = 3)) +
      scale_colour_manual(name = "High/Low Fibrosis Lake", 
                        values = c("#009E73", "grey"))+
   theme(legend.position = "None", 
        axis.text.y = element_text(size = 15),
        axis.text.x = element_text(size = 15),
        axis.title.x = element_text(size = 17),
        axis.title.y = element_text(size = 17, angle =90))+

      labs( x = "Infection prevalence (lake)",
            y = "Mean max tapeworm weight (g) per lake")

```

#Lake-level quick viz. of P:H biomass, find a visualization deep-dive in the Bayesian modeling portion below <3 

```{r PARASITE:HOST BIOMASS}
#how about just worm mass
cohort_data_full%>%
  ggplot(aes(x = Month.adj, 
             y = wormbiomass_fishmass, 
             #wormbiomass_fishmass = TREND HOLDS
             #total_worm_wt = TREND SWAMPS OUT
             #max_worm_wt = ZEROS SWAMP OUT
             #log(total_worm_wt+1) = TREND HOLDS
             #log(max_worm_wt+1)= ZEROS SWAMP OUT
             colour = Lake
             )) + 
  geom_jitter(width = 0.1, height = 0, 
              size = 4,
              alpha = 0.1) +
   stat_smooth(
            method = "gam", aes(group = 1),
            formula = y ~ s(x, k = 3),
            se = TRUE, colour = "black", fill = "grey50" ) +
  facet_wrap(~ Lake, ncol = 4) + 
  ylim(0,0.3)+
  labs(x = "", 
       y = "Lake-level parasite:host biomass fishes across time") +
  theme(axis.text.x = element_text(angle = 90), 
        legend.position = "None")

```

################################### 

# MODELS

################################### 

#gams
- bs = general purpose (standard continous)
- cc = cyclic cubic spline (endpoints joint smooths, good for season/months)
- cr/cs = cubric regression pline (continuous, numeric predictor; monotonic relationships)
- re = random effect (random intercept)
- fs = factor-smooth interaction (random slope, seperate smooth for each factor level, sharing a common penalty)
- ps  = pspline (great for dense, regular time series)
- mpi = shape-constrained additive model (SCAM) (monotone P-spline)

```{r MODELS: Fibrosis and Infection prevalence timing}
# Data for models
######################

fish_sizes_mod <- fish_sizes %>%
              filter(!Cohort == "F2") %>%
              mutate(fib_int = as.integer(Fibrosis) + 1, 
                     inf_yn = ifelse(Total_Worms>0, 1, 0)) %>% #for ocat model
              group_by(Lake, Month.adj) %>%
              mutate(inf_prev_lake_month = sum(inf_yn, na.rm=TRUE)/n()) %>%
              ungroup() %>%
              group_by(Lake, Month.adj, Cohort) %>%
              mutate(inf_prev_lake_month_cohort = sum(inf_yn, na.rm=TRUE)/n()) %>%
              ungroup()

cohort_data_mod <- cohort_data_full %>% #includes babies (but not their fibrosis)
                  filter(!Cohort == "F2") %>%
                  mutate(fib_int = as.integer(Fibrosis) + 1) #for ocat model

###############################
##################################
#ULTIMATE FIBROSIS GAM!##########
########## ADDING 
#Think about splines => HERE, s(Month_num) edf = ~2, means one broad curve (high in summer, low in winter)

#Wait! Let's also think about time-varying splines for fun.
#THIS IS THE MOST FABULOUS MODEL IN THE HISTORY OF EVER. 

m_fib_by <- mgcv::gam(fib_int ~    Cohort +  max_worm_wt* hi_low_fib +
                                  scale(inf_prev_lake_month_cohort) +
                                  s(Month_num, by = hi_low_fib, bs = "cc", k = 6) +
                                  s(Lake, bs = "re"),
                                  family = ocat(R = 5),
                                  data = fish_sizes_mod,
                                  knots = list(Month_num = c(1,13)),
                                  method = "REML")


#Mess around with level-order to get different comparisons in the model table
fish_sizes_mod$Cohort <- fct_relevel(fish_sizes_mod$Cohort, "F0") #"F0", "F1"

m_fib_2 <- mgcv::gam(
               fib_int ~ Cohort + max_worm_wt * hi_low_fib +inf_prev_lake_month_cohort+
                       # scale(inf_prev_lake_month_cohort) +
                          # deviation smooths by fibrosis regime
                          s(Month_num, by = interaction (hi_low_fib, Cohort), 
                            bs = "cc", k = 7) +
                          
                          s(Lake, bs = "re"),
                          family = ocat(R = 5),
                          data   = fish_sizes_mod,
                          knots  = list(Month_num = c(1, 13)),
                          method = "REML")

summary(m_fib_2)

AIC(m_fib_2, m_fib_by) #looks like that interaction is helpful!

simulationOutput <- simulateResiduals(fittedModel = m_fib_2, plot = T)
DHARMa::testDispersion(m_fib_2)
DHARMa::testZeroInflation(m_fib_2)


#Let's generate a table
tab_model(
  m_fib_2,
  show.ci = FALSE,
  show.se = TRUE,
  show.stat = TRUE,
  show.icc = FALSE,
  transform = NULL,
  dv.labels = "Fibrosis score (ordinal GAM)",
  pred.labels = c(
    "Intercept",
    "Cohort [F1]",
    "Max tapeworm weight (g) in fish",
    "Fibrosis regime [Low]",
    "Scaled infection prevalence",
    "Max worm weight × Fibrosis regime",
    # threshold parameters and smooths will auto-label
    "s(Month) [High Fib × F0]",
    "s(Month) [Low Fib × F0]",
    "s(Month) [High Fib × F1]",
    "s(Month) [Low Fib × F1]",
    "Lake (RE)" ))

#############################################
#Now it's time to simulate these maxima!!! 
# The question is WHEN DOES FIBROSIS peak? *(based on model estimates)
########################################

# 1) Build newdata with proper factor levels + a valid Lake
nd <- expand.grid(
        Month_num = seq(1, 13, by = 0.1), #get all the months in there
        hi_low_fib = levels(fish_sizes_mod$hi_low_fib),
        Cohort = levels(fish_sizes_mod$Cohort)[1:2])  # e.g., [1] for 1; "F1" or use both later

#Noq
nd <- nd %>%
  mutate( max_worm_wt = median(fish_sizes_mod$max_worm_wt, na.rm = TRUE), 
    #hold the parasite:host ratio at a typical value so that the time pattern is not confounded
    
    inf_prev_lake_month_cohort = mean(fish_sizes_mod$inf_prev_lake_month_cohort, na.rm = TRUE),
    # ibid
    
    Lake = factor(levels(fish_sizes_mod$Lake)[1], levels = levels(fish_sizes_mod$Lake)),
    # give a real lake level so mgcv's predict has a real valid factor level to work with. 
    #We DONT actually use this, it's just a technical requirement from this 
    #predict gam (see how we do exlude Lake later on)
    
    hi_low_fib = factor(hi_low_fib, levels = levels(fish_sizes_mod$hi_low_fib)),
    #reapply factor levels to match the fitted model
    Cohort     = factor(Cohort,     levels = levels(fish_sizes_mod$Cohort)) )

# 2) Predict category probabilities (R = 5) at population level
P <- predict(m_fib_2, newdata = nd, type = "response", exclude = "s(Lake)")

# 3) Expected fibrosis (0..4) and Pr(fibrosis ≥ 2)
k <- 0:4.  ##*1-5?
nd$E_fibrosis <- as.vector(P %*% k) ## Expected firbsosi score at each time point expected value E[Y] = sum_k k * Pr(Y = k)

# 4) Peak months (use either E_fibrosis or Pr_ge2)
peaks_E <- nd %>% group_by(hi_low_fib) %>%
  slice_max(E_fibrosis, n = 1, with_ties = FALSE) %>%
  transmute(hi_low_fib, peak_month = Month_num, peak_value = E_fibrosis)

peaks_E   # expected-score peak by lake type

# 5)  plot the line + dashed peak
ggplot(nd, aes(Month_num, E_fibrosis, colour = Cohort)) +
  geom_line(linewidth = 1.1) +
  geom_vline(data = peaks_E, aes(xintercept = peak_month, colour = hi_low_fib),
             linetype = "dashed") +
  facet_wrap(~hi_low_fib)+
  labs(y = "Expected fibrosis (0–4)", x = "Month") 

```

Here, we are showing how fibrosis and infection prevalence aren't synched.
```{r Correlation between infection and fibrosis}
#Let's summarize the data
lake_month <- fish_sizes_mod %>%
  group_by(Lake, Month.adj) %>%
  summarise(
    mean_fib = mean(Fibrosis, na.rm = TRUE),
    prev = mean(inf_prev_lake_month, na.rm = TRUE),
    .groups = "drop" )

#Lake specific correlations with bootstrapped CIs
cor_boot <- function(data, indices) {
  d <- data[indices, ]
  return(cor(d$prev, d$mean_fib))
}

cor_results <- lake_month %>%
  group_by(Lake) %>%
  #computes tibble for eahc lake and binds them together! never used that before, cool!
  group_modify(~{
    if(nrow(.x) < 4) return(tibble(cor = NA, lower = NA, upper = NA))
    b <- boot(.x, statistic = cor_boot, R = 2000) #Resample rows with replacement 2000times
    tibble(
      cor = mean(b$t, na.rm = TRUE),
      lower = quantile(b$t, 0.025, na.rm = TRUE), #ncie bootstrapped CIs
      upper = quantile(b$t, 0.975, na.rm = TRUE)
    )
  })

cor_results

#Let's plot 'em out bb!
ggplot(cor_results, aes(x = cor, y = reorder(Lake, cor))) +
  geom_vline(xintercept = 0, color = "grey60", linetype = "dashed") +
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0.2, color = "grey40") +
  geom_point(size = 5, color = "black") +
  labs(
    x = "Correlation between infection prevalence and fibrosis",
    y = "Lake") +
  theme_bw() +
   theme(
        axis.text.x = element_text(size = 14),
        axis.title.x = element_text(size = 18),
        axis.text.y = element_text(size = 14),
        axis.title.y = element_text(size = 18))

###

```

Here, I want to demonstrate that transmissibility of worms (>0.05g) varies between lakes, time, and stages.
Let's take the maximum worm weight in each fish and see if it is of transmissible stage.

```{r MODELS: Transmissible parasites}

#Do low-fibrosis lakes have more transmissible parasites than high-fibrosis lakes?

trans<- cohort_data_full %>%
        mutate(trans_yn = ifelse(max_worm_wt >= 0.05, 1, 0)) %>%
        select(Lake, Fibrosis, Total_Worms, max_worm_wt, hi_low_fib, trans_yn)

kruskal.test(trans_yn ~ hi_low_fib, data = trans)

#Fisher's Exact Test
table_df <- table(trans$hi_low_fib, trans$trans_yn)
table_df
fisher.test(table_df)

```

In a beta regression, phi controls precision. 

(1) High phi == low variability == consistent infection outcomes.
(2) Fibrosis negatively affects parasite biomass ratio.
(3) Fibrosis effect differs by lake type

Read this: https://bookdown.org/content/4857/monsters-and-mixtures.html

```{r BAYESIAN MODEL: Full zero-inf beta : PH: Fibrosis : Time }
#this is the one we are going for
bf_zib_point <- bf(
              wormbiomass_fishmass ~
                    #mu set up
                    fib_int * hi_low_fib + Cohort + (1 | Lake)+
                    s(Month_num, by = interaction(hi_low_fib, Cohort), bs = "cc", k = 5) ,
                    #zero inflated
                    zi  ~  fib_int * hi_low_fib + Cohort + (1 | Lake) +
                        s(Month_num, by = interaction(hi_low_fib, Cohort), 
                                     bs = "cc", k = 4) ,
                    #precision (beta)
                    phi ~ fib_int * Cohort + hi_low_fib + (1 | Lake) + #changed + for *
                          s(Month_num, by = interaction(hi_low_fib, Cohort), 
                                      bs = "cc", k = 3))


#Okay, maybe I can go a bit simpler?
#Removing the interaction is a no-go.
bf_simple <- bf( wormbiomass_fishmass ~
                  fib_int * hi_low_fib + Cohort + (1 | Lake) +
                  s(Month_num, by = interaction(hi_low_fib, Cohort), bs = "cc", k = 5),
                zi  ~ fib_int * hi_low_fib + Cohort + (1 | Lake),
                phi ~ fib_int * Cohort + hi_low_fib + (1 | Lake))


#######################################################
pri_zib <- c(
  # mean (mu) model
  prior(student_t(3, 0, 2.5), class = "Intercept"),
  prior(normal(0, 0.6), class = "b"),
  #prior(exponential(1), class = "sds"), #remove for simple model
  prior(student_t(3, 0, 2.5), class = "sd"),

  # zero-inflation (zi) model
  prior(student_t(3, 0, 2.5), class = "Intercept", dpar = "zi"),
  prior(normal(0, 0.6), class = "b",   dpar = "zi"),
 # prior(exponential(1), class = "sds", dpar = "zi"),  #remove for simple model
  prior(student_t(3, 0, 2.5), class = "sd",  dpar = "zi"),

  # precision (phi) model  (phi link is log by default)
  prior(normal(log(40), 0.6), class = "Intercept", dpar = "phi"),
  prior(normal(0, 0.3),       class = "b",         dpar = "phi"),
 # prior(exponential(1),       class = "sds",       dpar = "phi"),  #remove for simple model
  prior(student_t(3, 0, 2.5), class = "sd",        dpar = "phi"))

#######################################################

# !!!! DO NOT RUN THIS UNLESS YOU WANT TO WAIT AN HOUR !!! 
#More complex, lowest loo
fit_zib_point<- brms::brm(
                    bf_zib_point, data = inf_moddf,
                    family = brms::zero_inflated_beta(),
                    prior  = pri_zib,  # reuse your priors
                    knots  = list(Month_num = c(1, 13)),
                    chains = 4, iter = 3000, warmup = 1000,
                    threads = threading(2),
                    control = list(adapt_delta = 0.99, max_treedepth = 14),
                    seed = 2025)

# !!!! DO NOT RUN THIS UNLESS YOU WANT TO WAIT AN HOUR !!! 
#simpler with the same insight as fit_zib_point, thus will be our ultimate choice
fit_simple <- brms::brm(
                    bf_simple, data = cohort_data_mod , 
                    #includes babies, but not their fib scores.
                    family = brms::zero_inflated_beta(),
                    prior  = pri_zib,  # reuse your priors, but remove the SDS.
                    knots  = list(Month_num = c(1, 13)),
                    chains = 4, iter = 3000, warmup = 1000,
                    threads = threading(2),
                    control = list(adapt_delta = 0.99, max_treedepth = 14),
                    seed = 2025)


#######################################################################
## MODEL SELECTION
loo_zib_point <- loo(fit_zib_point)
loo_simp <- loo(fit_simple)

loo_compare(loo_simp, loo_zib_point) #meh, simpler is easier to interpret and results are the same + dharma is better for simple model

#### OUTPUT ####

summary(fit_simple) #fit_zib_point, fit_zib_refined
bayesplot::mcmc_intervals(fit_simple, regex_pars = "b_") +theme_classic()
bayesplot::mcmc_intervals(fit_simple, regex_pars = "^sds_") +theme_classic()

#################
#### QUALITY CHECKS 
#################
plot(fit_simple) #fit_zib_point, fit_zib_refined
neff_ratio(fit_simple) #→ all > 0.1 (preferably > 0.4) 

pp_check(fit_simple, type = "stat", stat = "mean") 


y    <- fit_simple$data$wormbiomass_fishmass
yrep <- posterior_predict(fit_simple, ndraws = 500) 

# correct PPC for a scalar statistic
ppc_stat(y, yrep, stat = function(z) mean(z == 0)) +
  labs(title = "P(Y = 0) overall")

# Observed proportion of zeros
obs_zero <- mean(y == 0)
# Simulated proportion of zeros for each posterior draw
rep_zero <- apply(yrep, 1, function(x) mean(x == 0))

hist(rep_zero, breaks = 30,
     main = "Posterior predictive check for zero inflation",
     xlab = "Proportion of zero predictions") +
     abline(v = obs_zero, col = "red", lwd = 3)


#####################
#DHARMA QUALITY CHECK!
#####################

# 1) Extract the exact data used inside the fit (after NA dropping etc.)
mf <- model.frame(fit_simple)  # same row order/rows as posterior_predict columns
y  <- mf$wormbiomass_fishmass

# 2) Posterior draws
yrep     <- posterior_predict(fit_simple, ndraws = 1000)  # draws x N_used
mu_draws <- posterior_epred(fit_simple, ndraws = 1000)    # draws x N_used

# 3) DHARMa wants N_used x draws
sim_mat <- t(yrep)                       # N_used x draws
mu_hat  <- colMeans(t(mu_draws))         # length N_used


# 4) Build DHARMa object
model.check <- createDHARMa(
  simulatedResponse       = sim_mat,
  observedResponse        = y,
  fittedPredictedResponse = mu_hat,
  integerResponse         = FALSE)

DHARMa::plotQQunif(model.check)            # QQ/uniform plot

# Extract the exact data used in the brms model (after NA filtering)
fitted_vals <- fitted(fit_simple, summary = TRUE)[, "Estimate"] #fit_zib_point

# Check alignment
length(fitted_vals) == model.check$nObs  # should be TRUE

# Now plot residuals vs fitted

#We have a slighhht deviation on an upper QT band: BUT. This is very common when the fitted predictor spans a small 0–0.04 range. The spline is already absorbing the major temporal pattern, and the small red wiggle likely reflects residual heterogeneity at high fitted values (peak months in low fibrosis lakes).
plotResiduals( model.check, form = fitted_vals,quantreg = TRUE)

# residual vs predicted values
plotResiduals(model.check, form = model.frame(fit_simple)$hi_low_fib) 


#####################
# TEMPORAL AUTOCORRELATION
#####################
library(spdep)

resid <- residuals(model.check)  #residuals from this is from the dharma quality check
#Month infomation
month <- mf$Month_num
mf <- model.frame(fit_simple)

# Aggregate residuals by month
df_resid <- data.frame(month = month, resid = resid)
resid_month <- aggregate(resid ~ month, data = df_resid, FUN = mean)

# Build 2D coordinates for 1D time
coords <- cbind(resid_month$month, rep(0, nrow(resid_month)))

# Nearest-neighbor weights
nb <- knn2nb(knearneigh(coords, k = 1))
lw <- nb2listw(nb, style = "W")

# Moran's I on temporal structure
moran.test(resid_month$resid, lw)


```

```{r BAYESIAN VISUALIZATION}

#Let's see what these splines are trying to tell us
##############
## GOING HARD! #
#################

# 1) Work on an F0/F1-only copy, with levels reset
inf_F01 <- cohort_data_mod %>%
  filter(Cohort %in% c("F0","F1")) %>%
  mutate( Cohort     = factor(Cohort, levels = c("F0","F1")),
          hi_low_fib = factor(hi_low_fib, levels = levels(cohort_data_mod$hi_low_fib)))

# 2) Median fib_int per stratum (or swap in your weighted version)
fib_meds <- inf_F01 %>%
  group_by(hi_low_fib, Cohort) %>%
  #Here we aim for median because of the skewed distribution. We want to capture what's really going on in the entire group
  summarise(fib_int = median(fib_int, na.rm = TRUE), .groups = "drop") 

# 3) Prediction grid 
nd <- fib_meds %>%
  expand_grid(Month_num = seq(1, 13, length.out = 200))

##############################
### Let's graphically check it out #
#############################

#Peak magnitude  and timing for each group

#These are posterior draws from the expected value of the response, conditional on my predictors
#For a zero inflated beta model, the expected value is : E[Y] = (1 − zi) * mu
ep <- posterior_epred( fit_simple, newdata = nd, re_formula = NA)   # marginalize over lake RE

nd$group <- paste(nd$hi_low_fib, nd$Cohort, sep = "_")

#for each posterior draw, compute peak PH for each group
# extract unique groups
groups <- unique(nd$group)

# helper: indices for each group
group_idx <- lapply(groups, function(g) which(nd$group == g))
names(group_idx) <- groups

# number of posterior draws
nsamp <- dim(ep)[1]

# storage lists
peak_value <- matrix(NA, nrow = nsamp, ncol = length(groups))
peak_month <- matrix(NA, nrow = nsamp, ncol = length(groups))

for (j in seq_along(groups)) {
  idx <- group_idx[[j]]
  # subset epred draws to this group over the 200 time points
  sub_ep <- ep[, idx]  # draws x timepoints
  # find peak per draw
  peak_value[, j] <- apply(sub_ep, 1, max) #POSTERIOR DRAWS OF PEAK PH FOR EACH GROUP
  # which time index corresponds to peak?
  peak_time_idx <- apply(sub_ep, 1, which.max)
  # convert index back to Month_num
  peak_month[, j] <- nd$Month_num[idx][peak_time_idx] #POSTERIOR FRAW OF PEAK MONTH FOR EACH GROUP
}

colnames(peak_value) <- groups
colnames(peak_month) <- groups

summarize_post <- function(x) {
  tibble(
    mean  = mean(x),
    lower = quantile(x, 0.025),
    upper = quantile(x, 0.975)
  )
}

########## EXTRACTING THE VALUES FORM POSTERIOR DRAWS
peak_value_summary <- map_dfr(as.data.frame(peak_value), summarize_post, .id = "Group")

peak_month_summary <- map_dfr(as.data.frame(peak_month),summarize_post,.id = "Group")

peak_value_summary
peak_month_summary

# ---- PLOTS ----
# A) Overall expectation (zeros mixed in)

facet_labs <- c(Low = "Low fibrosis lake", High = "High fibrosis lake")

########## Overall expectation #############

nd %>%
  add_epred_draws(fit_simple, re_formula = NA, ndraws = 1000) %>%
  ggplot(aes(x = Month_num, y = .epred, colour = Cohort)) +
  facet_wrap(~hi_low_fib, 
             labeller = labeller(hi_low_fib = facet_labs))+
  stat_lineribbon(.width = c(0.05, 0.1, 0.25), alpha = 0.25,
                  linewidth = 1.4) + 
  scale_colour_manual(values = c("#FF6347", "#477187"),
                      name = "Cohort")+ #"#009E73", "grey"
  scale_fill_manual(values= c("#ACABA5", "#797975", "#444342"), 
                    name = "Posterior credible interval") +
  labs(x = "", 
       y = (E[Y])~"Expected parasite:host biomass ratio",) +
  scale_x_continuous( breaks = 1:13, labels = levels(cohort_data_mod$Month.adj)) +
  scale_y_continuous(breaks = seq(0, 0.04, by = 0.005))+
  theme(axis.title.y = element_text(size = 18, angle = 90),
        axis.text.y = element_text(size = 16),
        axis.title.x = element_text(size = 18),
        legend.position = "None",
        strip.text.x = element_text(margin = margin(0.25,0,0.25,0, "cm")))
  

############################################

# B) Zero probability: P(Y = 0)

p_zero <- 
nd %>%
  add_epred_draws(fit_simple, dpar = "zi", re_formula = NA, ndraws = 1000) %>%
  ggplot(aes(x = Month_num, y = .epred, colour = Cohort)) +
  stat_lineribbon(.width = c(0.05, 0.1, 0.5), alpha = 0.25, 
                  linewidth = 1.4) + #c(.5, .8, .95)
  facet_wrap(~ hi_low_fib, nrow = 1) +
  scale_fill_brewer(palette = "Greys") +
  labs(x = "Month", y = "P(Y = 0)") +
  scale_x_continuous(breaks = seq(1, 13, by = 1))+
  theme(axis.title.y = element_text(size = 18, angle = 90),
        axis.text.y = element_text(size = 16),
        axis.title.x = element_text(size = 18), 
        axis.text.x = element_text(size = 16))
p_zero

# C) Positive-part mean: E[Y | Y > 0]
p_mu <- 
nd %>%
  add_epred_draws(fit_simple, dpar = "mu", re_formula = NA, ndraws = 1000) %>%
  ggplot(aes(x = Month_num, y = .epred, colour = Cohort)) +
  stat_lineribbon(.width = c(0.05, 0.1, 0.5, 0.8 ), alpha = 0.25, 
                  linewidth = 1.4) + #c(.5, .8, .95)
  facet_wrap(~ hi_low_fib, nrow = 1) +
  scale_fill_brewer(palette = "Greys") +
  labs(x = "Month", y = "E[Y | Y > 0]") +
  scale_x_continuous(breaks = seq(1, 13, by = 1))+
  theme(axis.title.y = element_text(size= 18, angle = 90),
        axis.text.y = element_text(size = 16),
        axis.title.x = element_text(size = 18), 
        axis.text.x = element_text(size = 16))
p_mu


```



########## END ###########
