---
title: "Fibrosis_Apr2026"
author: "Chloe Allison Fouilloux"
date: "2026-03-31"
output: html_document
---

Hello, 

Welcome to our very cool, very awesome **immunity/infection** dataset. Here, we look at 8 lakes every month over the course of a year. We sample the "super"model, the threespine stickleback, looking for the presence and intensity of infection by the trophically transmitted tapeworm, *Schistocephalus solidus*. Paired with looking into infection dynamics in natural populations we also are able to get a pulse of immune dynamics within these same hosts by the extent of fibrosis present in the peritoneal cavity of fish. How cool is that!

This look a lot of time to wrangle and code, so I am very happy for everyone to look through everything and provide feedback and insight. Thank you for your light and support! 

Shine bright like a tapeworm. Hugz, C.

- Tetative Manuscript Title: *Immuno-genetic variation and host ontogeny constrain pathogen virulence and transmission in a helminth parasite*

- Authors: Chloe A. Fouilloux1*, Heather Alexander2, Grace Vaziri2, Patrick McNamara2, Emma Polard2, Amanda Hund3, Edith Reyes Contreras3, Penelope Steffen3, Cate Patterson3, Shira Dubin3, Amy Chen3, Gwen Casey3, John Berini III3, Emma Choi4, Noelle Plantier4, Dina Soklovskaya4,  Daniel I. Bolnick4, Jessica L. Hite1

```{r Load Libraries}

library(tidyverse)
library(tidyr)
library(dplyr)
library(ggplot2)
library(forcats)
library(gghighlight)
library(ggpubr)
library(lubridate)
library(ggridges)
library(mclust)
library(purrr)
library(mgcv)
library(cluster)
library(DHARMa)
library(brms)
library(boot)
library(scam)

library(spdep)
library(sjPlot)

library(tidybayes)
library(bayesplot)

```

Set theme for ggplot

```{r Themes}
#Theme set
theme_set(  theme(
                  strip.background = element_rect(fill = "White"),
                  panel.background = element_rect(fill = "white",
                                                  colour = "black"), 
                  panel.grid.major = element_blank(), 
                  panel.grid.minor = element_blank(), 
                  text = element_text(size = 15)))
```

```{r Loading Data}

#These are little fish analyzed at UConn. These little guys are important for cohort analysis and infection prev!
baby_fish <- read_csv("baby_fish.csv")

#These are fish from BAMF : Dissected by HA and should be used for fibrosis scores
fish_sizes <- read_csv("fish_sizes.csv")

#worm weight by fish
worm_weight<- read.csv("worm_weights.csv")%>%#merge it with the ‘worm_weight’ file.
          mutate(Lake = as.factor(Lake),
                 FishID = as.factor(Fish_ID))%>% 
                 drop_na(FishID) %>%
                mutate(total_worm_wt = rowSums(select(., contains("worm_wt_")), 
                                               na.rm = TRUE)) 

```

```{r Bird Data}

#Data
birds<- read.csv("piscivorous_calls.csv")
loons <- read.csv("loon_calls.csv")

#Piscivorous bird data
birds<- birds %>% mutate(Month = as.factor(month), 
                         Species = as.factor("All"))

birds$Month <- fct_relevel(birds$Month, c(   "August",  "September", "October", "November", "December",  "January", "February", "March", "April", "May", "June", "July"))

#Loons bird data
loons<- loons %>% mutate(Month = as.factor(month), 
                         Species = as.factor("Loon"))

loons$Month <- fct_relevel(birds$Month, c(  "August",  "September", "October", "November", "December",  "January", "February", "March", "April", "May",  "June", "July"))


bamf_birds <- full_join(birds, loons)

```


TIDYING DATA, please read the notes within!

```{r Tidying Data}
baby_fish <- baby_fish %>%
             mutate(Lake = as.factor(Lake), 
                    Season = as.factor(Season),
                    Month.adj = as.factor(Month.adj), 
                    JOIN_ID = as.factor(JOIN_ID)) %>%
             #If the fish has a tapeworm, but the weight was never recorded it is because it was too small. So it gets 0.00005g.
            mutate(total_worm_wt = ifelse(Total_Worms > 0 & is.na(total_worm_wt), 0.00005, total_worm_wt), 
                   total_worm_wt = ifelse(Total_Worms == 0, 0, total_worm_wt),
                   max_worm_wt = ifelse(Total_Worms > 0 & max_worm_wt == 0,  0.00005, max_worm_wt))

fish_sizes <- fish_sizes %>%
             mutate(Lake = as.factor(Lake), 
                    Season = as.factor(Season),
                    Month = as.factor(Month),
                    Month.adj = as.factor(Month.adj), 
                    JOIN_ID = as.factor(JOIN_ID), 
                    Sex = as.factor(Sex), 
                    trans_worm_yn = ifelse(max_worm_wt>= 0.05, 1, 0)) %>%
           #If the fish have at least 1 tapeworm, but tapeworm weight is NA, then it should be assigned
           # 0.00005 g
            mutate(total_worm_wt = ifelse(Total_Worms > 0 & is.na(total_worm_wt), 0.00005, total_worm_wt), 
                   max_worm_wt = ifelse(Total_Worms > 0 & max_worm_wt == 0,  0.00005, max_worm_wt))

#LET'S GET TIME ALL IN ORDER, SHALL WE?
fish_sizes$Month.adj <- forcats::fct_relevel(fish_sizes$Month.adj, 
                                         c( "JUL.23", "AUG", "SEP", "OCT", 
                                           "NOV", "DEC", "JAN", "FEB", 
                                           "MAR", "APR", "MAY", "JUN", "JUL.24"))

fish_sizes$Season <- forcats::fct_relevel(fish_sizes$Season, 
                                         c("Summer","Fall", 
                                           "Winter", "Spring"))


#Remember here that max_worm_weight for baby_fish is just the total worm weight. None of these weights are yet transmissible (even when summed), and they are just used for data visualization purposes. 

ALL_fish <- bind_rows(
            fish_sizes %>%
              mutate(Data = factor("adult_fish", levels = c("baby_fish", "adult_fish"))),
            
            baby_fish %>%
              mutate(Fibrosis = NA_real_,
                     Data = factor("baby_fish", levels = c("baby_fish", "adult_fish"))))

ALL_fish <- ALL_fish %>%
            mutate(intensity_lengthcorr = (Total_Worms/Length),
                   intensity_masscorr = (Total_Worms/Mass), 
                  wormbiomass_fishmass = total_worm_wt/Mass)

```


################################### 

# COHORTS

################################### 

1.  In July–Sept 2023, larger modal group = F0, smaller = F1.
2.  After winter (say, Feb–March), the F0s should start disappearing.
3.  By summer 2024, F1s have grown and a new small group (F2s) may
    appear.
4.  Fit a mixture model to each lake × month subset.
5.  In the first summer (July–Sept 2023), rank by mean length to label
    cohorts (F0 = largest group).

#First of all, lakes have vastly different sized fish
```{r Fish Sizes and Sampling Counts}

#How many fish did we sample every month?
sampling_effort <- ALL_fish %>%
                   group_by(Lake, Month.adj) %>%
                   summarize(n = n())


ALL_fish %>%
mutate(Lake = fct_reorder(Lake, Length, .desc = TRUE)) %>%
ggplot(aes(x = Lake, y = Length )) +
    geom_jitter(size = 3, alpha = 0.2, height = 0, width = 0.1)+
    stat_summary(geom = "pointrange",
                 shape = 21,
                 size = 1,
                 colour = "salmon",
                 fill = "white",
                 fun.data = mean_cl_boot) +
    ylab("Length (mm)")

#Fish sizes between lakes significantly differ from one another (which is one reason why we can't use simple size cutoffs)

#Kruskal-Wallis test here is for non-parametric data.
kruskal.test(Length ~ Lake, data = ALL_fish)

```

#MODEL-BASED CLUSTERING APPROACH BASED ON FINITE NORMAL MIXTURE
MODELLING.

Here we are doing model-based clustering based on fish length. **Using
mclust.** + Gaussian Finite Mixture Mdoels (GMM) + Subsetted early fish,
(G = 1:2) for F0/F1s. Mclust assigns based on BIC. + Assuming that fish
length data are drawn from a mixture of gaussian distrubutions. + Aimed
at discovering inherent groups within data without prior knowledge of
labels

Read more here:
<https://luca-scr.r-universe.dev/mclust/doc/manual.html#Mclust>

GMMs = assumes each fish length (xi) comes from one of G normal
components; Mixture model below where: 

\pi_k = mixture proportion (the fraction of fish in cohort K, relative abundance); this is the prior probability that a random observation belongs to cohort K **(before seeing the data!)**

\mu_k = mean length of cohort k

\sigma_k\^{2} = variance within that cohort

$$
p(x) \;=\; \sum_{k=1}^{G} \pi_k \,\mathcal{N}\!\bigl(x \mid \mu_k, \sigma_k^{2}\bigr),
\quad
\sum_{k=1}^{G}\pi_k = 1,\; \pi_k \ge 0.
$$

Mclust estimates all these parameters by maximizing the likelihood of our observed data. Then, for each fish, it computes the posterior probability and assignes the fish to whichever K gives the highest probability.

Here, the numerator represents the joint probability that the data point came from cohort K and has a value of x (PRIOR). The denominator sums those joint probabilities across all possible cohorts (PRIOR: here subscript 'j' instead of 'k' because it runs over all cohorts in the denominator), turning the ratio into a posterior probability (Bayes' theorem).

$$
P(z=k \mid x)
= \frac{\pi_k \,\mathcal{N}\!\bigl(x \mid \mu_k, \sigma_k^{2}\bigr)}
       {\sum_{j=1}^{G} \pi_j \,\mathcal{N}\!\bigl(x \mid \mu_j, \sigma_j^{2}\bigr)}.
$$

We will also be leveraging freely available data about stickleback growth rates from DEB database: https://www.bio.vu.nl/thb/deb/deblab/add_my_pet/entries_web/Gasterosteus_aculeatus/Gasterosteus_aculeatus_res.html 

[WrigWoot2004] Wright, H. A., Wootton, R. J., and Barber, I. (2004). Interpopulation variation in early growth of threespine sticklebacks (Gasterosteus aculeatus) under laboratory conditions. Can. J. Fish. Aquat. Sci., 61:1832--1838.

Wright :: SL = 0.906 * TL + 0.200, r^2 = 0.996 

```{r heck yeah this looks awesome!}

assign_cohorts_bio <- function(ALL_fish_df,
                                f2_months   = 12:13, ## babies in month 12–13, not before!
                                f2_cap_mm   = 25,    #Babies are not bigger than this at 2 months in age 
                               
                                #**** base F1 caps from DEB curve (mm), Month 1..13 
                                f1_cap_mm   = c(30,34,37,40,42,44,45,46,47,48,49,49,49),
                                # lake-specific bumps in mm for selected months
                                #SRA has super big fishies!
                                lake_cap_bump = list(SRA = list(months = 10:13, bump_mm = 20),
                                                     RSS = list(months = 10:13, bump_mm = 4))) {
#Sanity checks on inputs
  stopifnot(all(c("Lake","Month_num","Length","JOIN_ID") %in% names(ALL_fish_df)))
  stopifnot(length(f1_cap_mm) >= max(ALL_fish_df$Month_num, na.rm = TRUE))

#Quick data clean to make sure that everything is the right format and not INFs or NAs
  df <- ALL_fish_df %>%
          mutate(Month_num = as.integer(Month_num), Length = as.numeric(Length)) %>%
          filter(is.finite(Length), is.finite(Month_num))

# ---------- helper to fit a mixture on a single numeric vector ----------
  # helper: up to G comps, sorted means and class ids
  fit_components <- function(x, Gwant) {
    x <- x[is.finite(x)]    # drop non-finite values
    n <- length(x)
    if (n == 0) return(list(means = numeric(0), class = integer(0)))

# number of components cannot exceed number of unique values
    G <- max(1, min(Gwant, n, length(unique(x))))

# try a Gaussian mixture fit with up to G components
# Mclust selects model and G by BIC within the provided range (1:G), where G = a cohort! (seen as a population) 
#GAUSSIAN MIXTURE MODELS: if that throws an error (too few points)s, return a null instead of breaking my beautiful loop
    
    fit <- tryCatch({ if (G == 1) Mclust(x, G = 1) else Mclust(x, G = 1:G) }, error = function(e) NULL)

   # if mixture fit succeeded and has means
    if (!is.null(fit) && !is.null(fit$parameters$mean)) {
      mu  <- as.numeric(fit$parameters$mean) #mean length of cohort K
      ord <- order(mu)              #Order cohort by mean (small to large)
      mu  <- mu[ord]                #sorted means from small to large
  
  # fit$classification are component labels in Mclust’s internal order
  # we remap them into 1..G according to ascending mean so “1 = smallest”
      cl  <- match(fit$classification, ord) 
      #this is a maximum a posteriori (MAP) label = cohort with highest prob. for each fish
      return(list(means = mu, class = cl))
    }

#FOR WINTER MONTHS, DATA DEFICIENT LAKES: Simpler clustering approach, kmeans. Does not model variance or mixture weights
# fallback: kmeans if mixture fails; simpler geometric clustering
    km  <- kmeans(x, centers = G, nstart = 20)
    ord <- order(as.numeric(km$centers))
    mu  <- sort(as.numeric(km$centers))
    cl  <- match(km$cluster, ord)
    list(means = mu, class = cl)
  }
  
    # ---------- end helper ----------

    # ---------- 1) per lake × month: fit and label by rank ----------

  # 1) per lake x month fit + provisional label by rank
  by_month <- df %>%
    group_by(Lake, Month_num) %>%
    group_modify(function(d, key) { # called once per lake-month group
     
       # allow 3 comps only in late months where F2 may appear
      Gwant <- if (key$Month_num %in% f2_months) 3 else 2
      
      fm <- fit_components(d$Length, Gwant) # mixture fit on that month’s lengths
      
      # if no fit (empty group), return placeholders
      if (length(fm$class) == 0)
        return(d %>% mutate(Cohort0 = NA_character_, comp_mean = NA_real_))

      d$comp      <- fm$class
      mu_sorted   <- fm$means
      d$comp_mean <- mu_sorted[d$comp]

      # map components to cohorts based on size rank
      # months 1–11: small → F1, large → F0
      # months 12–13: small → F2, middle → F1, large → F0
      if (length(mu_sorted) == 1) {
        lab_map <- if (key$Month_num <= 3) c("F0") else c("F1")
      } else if (length(mu_sorted) == 2) {
        lab_map <- c("F1","F0")
      } else {
        lab_map <- c("F2","F1","F0")
      }
      d$Cohort0 <- lab_map[d$comp]
      d
    }) %>%
    ungroup()
  
   # ---------- 2) compute monthly cohort targets (means) ----------
  # for each lake-month-cohort, use the mean Length as the target

  # 2) month targets from means
  cohort_means <- by_month %>%
    filter(!is.na(Cohort0)) %>%
    group_by(Lake, Month_num, Cohort0) %>%
    summarise(mu = mean(Length), .groups = "drop") %>%

    # drop any F2 outside late months (safety)
    mutate(Cohort0 = ifelse(Cohort0 == "F2" & !(Month_num %in% f2_months), NA, Cohort0)) %>%
    filter(!is.na(Cohort0))

  # pivot to wide so we have columns F0, F1, F2 with their monthly mean targets
  targets <- cohort_means %>%
    transmute(Lake, Month_num, Cohort = Cohort0, mu_mon = mu) %>%
    tidyr::pivot_wider(names_from = Cohort, values_from = mu_mon)

    # ---------- 3) assign each fish to the nearest target ----------

  # 3) assign to nearest target
  assigned <- df %>%
    left_join(targets, by = c("Lake","Month_num")) %>%
    mutate(
      dF0 = ifelse(is.finite(F0), abs(Length - F0), Inf),
      dF1 = ifelse(is.finite(F1), abs(Length - F1), Inf),
      dF2 = ifelse(is.finite(F2), abs(Length - F2), Inf),
  #this looks at each row, and compares real fish length to the mean of each cluster. 
  #whichever cluster it is closest to, that's the one it will be assigned.
      Cohort = c("F0","F1","F2")[apply(cbind(dF0, dF1, dF2), 1, which.min)]
    )

    # ---------- 4) enforce biological caps (per month, with lake bumps) ----------

  # 4) apply biological caps with optional lake bumps
  # build a per-row cap for F1
  f1_cap_vec <- f1_cap_mm[pmin(assigned$Month_num, length(f1_cap_mm))]

  # apply lake-specific bumps
    # apply lake-specific bumps: for listed lakes and months, add bump_mm to the cap

  if (length(lake_cap_bump)) {
    for (lk in names(lake_cap_bump)) {
      months <- lake_cap_bump[[lk]]$months
      bump   <- lake_cap_bump[[lk]]$bump_mm
      idx <- which(assigned$Lake == lk & assigned$Month_num %in% months)
      if (length(idx)) f1_cap_vec[idx] <- f1_cap_vec[idx] + bump
    }
  }

  assigned <- assigned %>%
    mutate(
      # F2 only in late months and very small
      Cohort = ifelse(Month_num %in% f2_months & Length <= f2_cap_mm, "F2", Cohort),
      Cohort = ifelse(!(Month_num %in% f2_months) & Cohort == "F2", "F1", Cohort),

      # F1 cannot exceed cap -> becomes F0
      Cohort = ifelse(Cohort == "F1" & Length > f1_cap_vec, "F0", Cohort),

      # F0 cannot be below the F1 cap -> becomes F1
      Cohort = ifelse(Cohort == "F0" & Length <= f1_cap_vec, "F1", Cohort),

      Cohort = factor(Cohort, levels = c("F0","F1","F2"))
    ) %>%
    select(JOIN_ID, Lake, Month_num, Length, Cohort)

  assigned
}


###################################################################
##################################################################
ALLfish_with_cohorts <- assign_cohorts_bio(ALL_fish, f2_months = 12:13, f2_cap_mm = 25)

ALL_fish <- ALL_fish %>%
            left_join(ALLfish_with_cohorts %>%
                        select(JOIN_ID, Cohort) %>% 
                        distinct(),
                          by = c("JOIN_ID")) %>%
            drop_na(Cohort)

cohort_data_full <- ALL_fish%>%
                    mutate(DEB = (100*Mass/(Length^3)), na.rm = T) %>%
                    mutate(inf_yn = ifelse(Total_Worms>0, 1, 0), 
                           trans_yn = as.integer(max_worm_wt > 0.05)) %>% #a little broader ,# 0.045
                    group_by(Lake) %>% #X LAKE 
                    mutate(
                      inf_prev_lake = sum(inf_yn, na.rm=TRUE)/n(), 
                      inf_prev_lake_trans = sum(trans_yn, na.rm=TRUE)/n())  %>%
                    group_by(Lake, Month.adj) %>% #X LAKE X MONTH
                    mutate(
                      inf_prev_lake_month = sum(inf_yn, na.rm=TRUE)/n(),
                      inf_prev_lake_month_trans = sum(trans_yn, na.rm=TRUE)/n()) %>%
                    group_by(Lake, Cohort) %>% # XLAKE X COHORT
                    mutate(
                      inf_prev_lake_cohort = sum(inf_yn, na.rm=TRUE)/n()) %>%
                    ungroup() %>%
                    group_by(Lake, Month.adj, Cohort) %>%  #X LAKE X MONTH X COHORT
                    mutate(
                      inf_prev_lake_month_cohort = sum(inf_yn, na.rm=TRUE)/n()) %>%
                    ungroup() %>%
                    mutate(Season = as.factor(Season)) %>%
                    mutate(hi_low_fib = case_when(
                                  Lake %in% c("SUG", "SRA", "L2B") ~ "High",
                                  Lake %in% c("PCH", "FRE", "BLA", "RSS", "BEW")~ "Low",
                                 TRUE ~ NA_character_), 
                            hi_low_fib = as.factor(hi_low_fib)) 


fish_sizes <- fish_sizes %>%
              left_join(
                cohort_data_full %>% select(JOIN_ID, Cohort) %>% distinct(),
                by = c("JOIN_ID"))

####### Let's look at this ABSOLUTE BEAUTY OF A PLOT! ####
cohort_data_full %>%
mutate(Lake = fct_reorder(Lake, Fibrosis, .desc = TRUE)) %>%
ggplot(aes(x = Month_num, y = Length, color = Cohort)) +
  geom_point(alpha = 0.4, size = 4) +
#The smoothed line is generated using a shape-constrained additive model (SCAM), where the length–month relationship #is fit with a monotone P-spline (bs = "mpi").
  geom_smooth(method = scam::scam, #NOTE! this is from a seperate package!
              formula = y ~ s(x, k = 4, bs = "mpi"),
              se = T, linewidth = 1.2) +
  facet_wrap(~Lake, ncol = 4) +
  scale_x_continuous(
    breaks = unique(cohort_data_full$Month_num),
    labels = unique(cohort_data_full$Month.adj)) +
  labs(x = "", y = "Length (mm)") +
  scale_colour_manual(values = c("F0" = "#FF6347", 
                                "F1" = "#477187", 
                                "F2" = "lightblue")) +
theme(   strip.text.x = element_text(size = 16, 
                                     margin = margin(0.25,0,0.25,0, "cm")),
            axis.title.y = element_text(size = 22, 
                                        angle = 90, margin = margin(r = 10)),
            axis.title.x = element_text(size = 22, margin = margin(5)),
            axis.text.y = element_text(size = 18),
            #legend.position = "none" ,
            axis.text.x = element_text(size = 18) )  



```


################################### 

# Generating clusters of high/low fibrosis lakes

################################### 


```{r Hi-Low Lake Clustering (not int.prev.)}

lake_summary <- fish_sizes %>%
                group_by(Lake) %>%
                summarise(
                  fib_median   = median(Fibrosis, na.rm = TRUE),
                  fib_mean     = mean(Fibrosis, na.rm = TRUE),
                  fib_se =  sd(Fibrosis, na.rm = TRUE)/sqrt(sum(!is.na(Fibrosis))),
                
                  prp_trans = mean(trans_worm_yn),
                  se_trans = sqrt(prp_trans * (1 - prp_trans) / n()),
                  
                  inf_prev     = mean(Total_Worms > 0, na.rm = TRUE),
                  mean_worm_intensity = mean(Total_Worms, na.rm = TRUE),
                  mean_worm_burden = mean(total_worm_wt, na.rm = TRUE),
                  burden_se =  sd(total_worm_wt, na.rm = TRUE)/sqrt(sum(!is.na(total_worm_wt))),
                  .groups = "drop" )

lake_summary
# Then feed fib_median or fib_mean into PAM
#this is giving the highest weight to fibrosis

w <- c(fib_mean = 0.7,  inf_prev = 0.15, mean_worm_burden = 0.15)

#Daily is a dissimilarity matrix for mixed-type data
diss <- daisy( lake_summary %>% select(fib_mean,  inf_prev, mean_worm_burden),
               metric = "gower", stand = TRUE, weights = as.numeric(w))
diss

##################
### How many clusters should I choose for my PAM matrix?

sil_widths <- sapply(2:5, function(k) {
  pam(diss, k = k)$silinfo$avg.width
})
plot(2:5, sil_widths, type = "b", xlab = "k", ylab = "Average silhouette width")

######################

pam_fit <- pam(diss, k = 2)


lake_summary %>% mutate(cluster = factor(pam_fit$clustering))

###

###
fish_sizes <- fish_sizes %>% mutate(hi_low_fib = case_when(
                                                Lake %in% c("SUG", "SRA", "L2B") ~ "High",
                                                Lake %in% c("PCH", "FRE", "BLA", "RSS","BEW")~ "Low",
                                               TRUE ~ NA_character_), 
                                    hi_low_fib = as.factor(hi_low_fib))

ALL_fish <- ALL_fish %>% mutate(hi_low_fib = case_when(
                                                Lake %in% c("SUG", "SRA", "L2B") ~ "High",
                                                Lake %in% c("PCH", "FRE", "BLA", "RSS","BEW")~ "Low",
                                               TRUE ~ NA_character_), 
                                    hi_low_fib = as.factor(hi_low_fib))

```


################################### 
# VISUALIZATIONS
###################################

Here each chunk acts as a chapter of information, open up to unravel the fun!
As follows:
1. Fibrosis score + Temporal Viz 
2. Parsite burden (frequentist approach!) + Temporal Viz. (Note! We use Bayesian version for final plot)
3. Transmissible stages + Temportal Viz.
4. Definitive host occurence + Temporal Viz

```{r Fibrosis Score/Time}

#With a smoother
#FIBROSIS BY COHORT + HI/LOW FIBROSIS
A1<- fish_sizes %>%
  filter(!Cohort %in% c(NA ,"F2")) %>%
  ggplot(aes(x = Month.adj, colour =  hi_low_fib, fill = hi_low_fib,
             y = Fibrosis)) +
  geom_jitter(width = 0.1, height = 0.08, 
             size = 3, alpha = 0.01) +
  stat_smooth(method = "gam", 
              aes(group = hi_low_fib), alpha = 0.1,
              formula = y ~ s(x, bs = "tp", k = 3), #cr, cc
              fullrange = T ) +
  labs(x = "", y = "Fibrosis Score (0-4)") +
  theme(strip.text = element_text(size = 18),
        axis.text.y = element_text(size = 18),
        axis.title.y = element_text(size = 22, 
                          angle =90, margin = margin(r = 15))) +
  theme(legend.position = "None")+
  scale_colour_manual( values = c("Low" = "darkgrey",
                                "High"  = "#1D9E75")) +
 scale_fill_manual( values = c("Low" = "darkgrey",
                                "High"  = "#1D9E75")) 

A1
####### Now incorporating cohort into it

B1<- fish_sizes %>%
  filter(!Cohort %in% c(NA ,"F2")) %>%
  ggplot(aes(x = Cohort, colour =  hi_low_fib,
             y = Fibrosis)) +
  geom_jitter(width = 0.1, height = 0.08, 
             size = 3, alpha = 0.02) +
    stat_summary(geom = "line",
               aes(group = hi_low_fib), linetype = "dashed",
                 shape = 21,  size = 0.8, fill = "white",
                 fun.data = mean_cl_boot) +
  stat_summary(geom = "pointrange",
                 shape = 21,  size = 1,
                 stroke = 2, fill = "white",
                 fun.data = mean_cl_boot) +
  labs(x = "", y = "Fibrosis Score (0-4)") +
  theme(strip.text = element_text(size = 18),
        axis.title.x = element_text(size = 18),
        axis.text.x = element_text(size = 18),
        axis.text.y = element_text(size = 18),
        axis.title.y = element_text(size = 18, 
                        angle =90, margin = margin(r = 15))) +
  theme(legend.position = "None")+
  scale_colour_manual(
  values = c("Low" = "darkgrey",
             "High"  = "#1D9E75"))

B1

```

```{r Parasite burden/Time}

#If you want to see anything, you have to filter out the 0s

#wormbiomass: fishmass
A2<- cohort_data_full %>%
  filter(!Cohort %in% c(NA ,"F2")) %>%
  filter(Total_Worms > 0) %>%
  ggplot(aes(x = Month.adj, colour =  hi_low_fib,
             y = wormbiomass_fishmass)) + 
  #total_worm_wt, log(total_worm_wt+1), total_worm_wt/Mass
  #max_worm_wt, log(total_worm_wt+1), max_worm_wt/Mass
  #wormbiomass_fishmass
  geom_jitter(width = 0, height = 0, 
             size = 3, alpha = 0.01) +
  stat_smooth(method = "gam", 
              aes(group = hi_low_fib),
              formula = y ~ s(x, bs = "tp", k = 6), #cr, cc
              fullrange = T ) +
  labs(x = "", y = "Parasite:host biomass ratio (worm g/host g)") + #Size corrected infection burden
  theme(strip.text = element_text(size = 18),
        axis.text.y = element_text(size = 18),
        axis.title.y = element_text(size = 22, 
                          angle =90, margin = margin(r = 15))) +
  theme(legend.position = "None")+
  coord_cartesian(ylim = c(0, 0.3))+
  scale_colour_manual(
  values = c("Low" = "darkgrey",
             "High"  = "#1D9E75"))


A2


cohort_data_full %>%
  filter(!Cohort %in% c(NA ,"F2")) %>%
  filter(Total_Worms > 0) %>%
  ggplot(aes(x = Month.adj, colour =  hi_low_fib,
             y = Length)) + #max_worm_wt
  #total_worm_wt, log(total_worm_wt+1), total_worm_wt/Mass
  #max_worm_wt, log(total_worm_wt+1), max_worm_wt/Mass
  #wormbiomass_fishmass
  geom_jitter(width = 0, height = 0, 
             size = 3, alpha = 0.01) +
  stat_smooth(method = "gam", 
              aes(group = hi_low_fib),
              formula = y ~ s(x, bs = "tp", k = 6), #cr, cc
              fullrange = T ) +
  labs(x = "", y = "Length in infected hosts (mm)") + 
  #Size corrected infection burden
  #Max. worm weight in host (g)
  theme(strip.text = element_text(size = 18),
        axis.text.y = element_text(size = 18),
        axis.title.y = element_text(size = 22, 
                          angle =90, margin = margin(r = 15))) +
  theme(legend.position = "None")+
  #coord_cartesian(ylim = c(0, 0.3))+
  scale_colour_manual(
  values = c("Low" = "darkgrey",
             "High"  = "#1D9E75"))




####### Now incorporating cohort into it

B2<- 
  fish_sizes %>%
  filter(!Cohort %in% c(NA ,"F2")) %>%
    filter(Total_Worms > 0) %>%
  ggplot(aes(x = Cohort, colour =  hi_low_fib,
             y = wormbiomass_fishmass)) +
  geom_jitter(width = 0.08, height = 0, 
             size = 3, alpha = 0.02) +
    stat_summary(geom = "line",
               aes(group = hi_low_fib), linetype = "dashed",
                 shape = 21,  size = 0.8, fill = "white",
                 fun.data = mean_cl_boot) +
  stat_summary(geom = "pointrange",
                 shape = 21,  size = 1,
                 stroke = 2, fill = "white",
                 fun.data = mean_cl_boot) +
    coord_cartesian(ylim = c(0, 0.3))+

  labs(x = "", y = "Parasite:host biomass ratio (worm g/host g)") +
  theme(strip.text = element_text(size = 18),
        axis.title.x = element_text(size = 18),
        axis.text.x = element_text(size = 18),
        axis.text.y = element_text(size = 18),
        axis.title.y = element_text(size = 18, 
                        angle =90, margin = margin(r = 15))) +
  theme(legend.position = "None")+
  scale_colour_manual(
  values = c("Low" = "darkgrey",
             "High"  = "#1D9E75"))


B2

```

```{r Max # Transmissible stages}
#max worm weight by hi/low fib by month
#count transmissible parasites (sum trans worm yn)/total parasites (sum total parasites)

trans <- cohort_data_full %>%
          select(Month.adj, Lake, hi_low_fib, JOIN_ID, Cohort,Season, Month_num,
                 Total_Worms, max_worm_wt) %>%
          mutate(max_worm_wt = ifelse(Total_Worms == 0, 0, max_worm_wt), 
                 trans_worm_yn = ifelse(max_worm_wt >= 0.05, 1, 0)) %>%
          group_by(hi_low_fib, Month.adj) %>%
          mutate(prp_trans_stages = sum(trans_worm_yn, na.rm = T)/sum(Total_Worms, na.rm = T), 
                 tot_trans_stages = sum(trans_worm_yn, na.rm = T))%>%
          ungroup()

trans_co <- trans %>%
             group_by(hi_low_fib, Cohort) %>%
             filter(!Cohort == "F2") %>%
             summarise(tot_trans_stages = sum(trans_worm_yn, na.rm = T))


############################
### MAX INFECTION ########
########################


A4<- trans %>%
  #   filter(hi_low_fib == "High") %>%
  ggplot(aes(x = Month.adj, colour =  hi_low_fib,
             y = log(tot_trans_stages+1))) + 
  geom_jitter(width = 0, height = 0, 
             size = 3, alpha = 0.01) +
  stat_smooth(method = "gam", 
              aes(group = hi_low_fib),
              formula = y ~ s(x, bs = "tp", k = 3), #cr, cc
              fullrange = T ) +
  labs(x = "", y = "log(# transmissible stages +1)") + #Size corrected infection burden
  theme(strip.text = element_text(size = 18),
        axis.text.y = element_text(size = 18),
        axis.title.y = element_text(size = 22, 
                          angle =90, margin = margin(r = 15))) +
  theme(legend.position = "None")+
  scale_colour_manual(
  values = c("Low" = "darkgrey",
             "High"  = "#1D9E75"))


A4

####### Now incorporating cohort into it

B4<- trans_co %>%
  ggplot(aes(x = Cohort, colour =  hi_low_fib,
             y = tot_trans_stages)) +

    stat_summary(geom = "line",
               aes(group = hi_low_fib), linetype = "dashed",
                 shape = 21,  size = 0.8, fill = "white",
                 fun.data = mean_cl_boot) +
  stat_summary(geom = "pointrange",
                 shape = 21,  size = 1,
                 stroke = 2, fill = "white",
                 fun.data = mean_cl_boot) +
  ylim(0, 110)+
  labs(x = "", y = "# transmissible stages/total parasites") +
  theme(strip.text = element_text(size = 18),
        axis.title.x = element_text(size = 18),
        axis.text.x = element_text(size = 18),
        axis.text.y = element_text(size = 18),
        axis.title.y = element_text(size = 18, 
                        angle =90, margin = margin(r = 15))) +
  theme(legend.position = "None")+
  scale_colour_manual(
  values = c("Low" = "darkgrey",
             "High"  = "#1D9E75"))
B4

```

```{r Definitive Host Occurence}

#Not lake specific, note month offset. Bird data observations started in August. 

birds$Month <- factor(birds$Month,
                      levels = c("", levels(birds$Month)))
levels(birds$Month)

BRD<- birds %>%
  ggplot(aes(x = Month,
             y = piscivore_calls)) + 
      geom_line(aes(group = 1), size = 1) +
    scale_x_discrete(drop = FALSE) +
  labs(y = "Piscivore calls (count)", x = "")+
  theme(strip.text = element_text(size = 18),
        axis.text.y = element_text(size = 18),
        axis.title.y = element_text(size = 22, 
                          angle =90, margin = margin(r = 15))) +
  theme(legend.position = "None")

BRD

SP<- birds %>%
  ggplot(aes(x = Month,
             y = n_species)) + 
      geom_line(aes(group = 1), size = 1) +
    scale_x_discrete(drop = FALSE) +
  labs(y = "Number of piscovorous species (count)", x = "")+
  theme(strip.text = element_text(size = 18),
        axis.text.y = element_text(size = 18),
        axis.title.y = element_text(size = 22, 
                          angle =90, margin = margin(r = 15))) +
  theme(legend.position = "None")

SP

ggsave("bird.png", BRD, height = 4, width = 6)

########################################
#### CO-OCCURRENCE ####################

trans_sum <- trans %>%
              group_by(hi_low_fib, Month.adj) %>%
              filter(!Month.adj == "JUL.23") %>% #there isn't these data for the birds
  #Before the filter = 14 HI/181 Low
  #After the filter = 13 HI/ 139 Low
              summarise(tot_trans_stages = first(tot_trans_stages), 
                        .groups = "drop") %>%
              ungroup() %>%
              mutate(Month = dplyr::recode(Month.adj,
                                           "AUG" = "August",
                                           "SEP" = "September",
                                           "OCT" = "October",
                                           "NOV" = "November",
                                           "DEC" = "December",
                                           "JAN" = "January",
                                           "FEB" = "February",
                                           "MAR" = "March",
                                           "APR" = "April",
                                           "MAY" = "May",
                                           "JUN" = "June",
                                           "JUL.24" = "July"))

sum(trans_sum$tot_trans_stages[trans_sum$hi_low_fib == "High"])
sum(trans_sum$tot_trans_stages[trans_sum$hi_low_fib == "Low"])

infbird<- full_join(trans_sum, birds, by = "Month")

#infsp<- 
  infbird %>%
  #filter(hi_low_fib == "High") %>%
  ggplot(aes(x = n_species, colour =  hi_low_fib,
             #piscivore_calls
             y = tot_trans_stages)) + 
  stat_smooth(method = "gam", 
              fill = "lightgrey",
              aes(group = hi_low_fib),
              formula = y ~ s(x, bs = "tp", k = 3),
              fullrange = T ) +
    geom_point(size = 3, shape = 16) +
  labs(x = "Number of piscovorous species (count)", y = "# transmissible stages/total parasites") + #
  theme(axis.text.x = element_text(size = 18),
        axis.text.y = element_text(size = 18),
        axis.title.x = element_text(size = 22),
        axis.title.y = element_text(size = 22, angle =90, margin = margin(r = 15))) +
  theme(legend.position = "None")+
  scale_colour_manual(
  values = c("Low" = "darkgrey",
             "High"  = "#1D9E75"))


#what if we scale the trans. so we can better see the relationship?
infsp<- infbird %>%
  group_by(hi_low_fib) %>%
  mutate(tot_trans_scaled = scale(tot_trans_stages)) %>%
  ggplot(aes(x = n_species,
             y = tot_trans_scaled,
             colour = hi_low_fib)) +
   stat_smooth(method = "gam", 
              fill = "#B0B0B001",
              alpha = 0.05,
              aes(group = hi_low_fib),
              formula = y ~ s(x, bs = "tp", k = 3),
              fullrange = T ) +
    geom_point(size = 3) +
    labs(x = "Number of piscovorous species (count)", 
         y = "# transmissible stages/total parasites") + 
  theme(axis.text.x = element_text(size = 18),
        axis.text.y = element_text(size = 18),
        axis.title.x = element_text(size = 22),
        axis.title.y = element_text(size = 22, angle =90, margin = margin(r = 15))) +
  theme(legend.position = "None")+
  scale_colour_manual(values = c("Low" = "darkgrey",
                                   "High"  = "#1D9E75"))

infsp

```

```{r All glued together for megaplot}
#Let's put it all together
library(cowplot)
library(patchwork)

time<- A2/A1/A4/SP #A2 here is frequentist, but we use bayesian one for paper
age<-  B2/B1/B4/infsp #B3

time
ggsave("time.png", time, height = 10, width = 3.5)
ggsave("age.png", age, height = 10, width = 3.5)

#### For presentations

A1+B1
A2+B2

A4 + B4
A4/SP
infsp

```

#VIRULENCE-TRANSMISSION TRADE-OFF
```{r EMPIRICAL DATA FOR V-T TRADE OFF}

#We want to know, mean transmissible tapeworm size per lake. That means that across all of the worm measures here, look for the ones that are >0.05g (~158 worms).

lake_trans_worms <- worm_weight %>%
                  pivot_longer(
                    cols = starts_with("worm_wt_"), names_to = "worm_id",
                    values_to = "worm_wt_g" ) %>%
                   filter(!is.na(worm_wt_g), worm_wt_g > 0.05) %>% 
  # transmissible threshold
                   group_by(Lake) %>%
                     #NOTE: THERE ARE ONLY 7 LAKES HERE, BEW HAS NO TRANS. WORMS
                   summarise(
                            max_trans_g = max(worm_wt_g),
                            mean_trans_g = mean(worm_wt_g),
                            se_trans_g   = sd(worm_wt_g) / sqrt(n()),
                            n_trans_worms = n(),
                            n_fish_with_trans = n_distinct(FishID),
                            .groups = "drop" )

########################################################################
lake_summary1 <- fish_sizes %>% #no individual worm data on baby fish
                mutate( trans_yn = as.integer(trans_worm_yn == 1) ) %>%
                group_by(Lake) %>%
                summarise(
                  # Max worm weight: unconditional mean (includes zeros)
                  mean_max_worm = mean(max_worm_wt, na.rm = TRUE),
                  se_maxworm    = sd(max_worm_wt, na.rm = TRUE) /
                    sqrt(sum(!is.na(max_worm_wt))),
              
                  # Max worm weight among infected only (exclude zeros)
                  mean_max_worm_inf = mean(ifelse(max_worm_wt > 0, max_worm_wt,
                                                  NA_real_), na.rm = TRUE),
                  se_maxworm_inf    = sd(ifelse(max_worm_wt > 0, max_worm_wt,
                                                NA_real_),  na.rm = TRUE) /
                                      sqrt(sum(!is.na(ifelse(max_worm_wt > 0,max_worm_wt,
                                                             NA_real_)))),
                  
                 #infected prevalence over entire year
                  inf_prev_year = mean(Total_Worms > 0, na.rm = TRUE),
                  .groups = "drop" )

#####################################
 ### Integrated prevalence ####
#####################################

yearly_df <- cohort_data_full %>%
              arrange(Lake, Month_num) %>%
              group_by(Lake, hi_low_fib) %>%
              mutate(
                dx = Month_num - lag(Month_num),
                trap_area = dx * (inf_prev_lake_month + lag(inf_prev_lake_month)) / 2 , 
                trap_area_trans = dx * (inf_prev_lake_month_trans +     
                                            lag(inf_prev_lake_month_trans)) / 2) %>%
             summarise( int_inf_prev = sum(trap_area, na.rm = TRUE))
##################################################
   
# Step 1: monthly prevalence per lake
monthly_prev <- fish_sizes %>%
                  group_by(Lake, Month.adj) %>%
                  summarise(  inf_prev_month = mean(Total_Worms > 0, na.rm = TRUE),
                    .groups = "drop")

# Step 2: max monthly prevalence per lake
max_monthly_prev <- monthly_prev %>%
                    group_by(Lake) %>%
                    summarise(  
                      max_inf_prev = max(inf_prev_month, na.rm = TRUE),
                      mean_inf_prev = mean(inf_prev_month, na.rm = TRUE),
                      .groups = "drop"  )
        
  
lake_summary1 <- lake_summary1 %>% left_join(yearly_df, by = "Lake")
lake_summary1 <- lake_summary1 %>% left_join(max_monthly_prev, by = "Lake")

#With trans as welll
vt<-
lake_summary1 %>%
  ggplot() + #inf_prev
  stat_smooth( aes(y = int_inf_prev, x = mean_max_worm_inf, group = 1), 
              method = "gam",
              fill = "lightgrey", colour = "black",
              formula = y ~ s(x, bs = "tp", k = 7)) +
          labs(y = "Integrated infection prevalence
(proportion × months)", 
               x = "Parasite burden 
(max. worm weight (g) per infected host)") +
        
      geom_text(aes(label = Lake, y = int_inf_prev, x = mean_max_worm_inf), vjust = -1.5, 
                hjust = 0.85, size = 3)  +
                # vertical error bars (SE on infection prevalence)
  geom_errorbarh(aes(xmin = mean_max_worm_inf - se_maxworm_inf,
                    xmax = mean_max_worm_inf + se_maxworm_inf, 
                    y = int_inf_prev, x = mean_max_worm_inf), width = 0) +
    geom_point(aes(y = int_inf_prev, x = mean_max_worm_inf, 
                 colour = hi_low_fib),  size = 4) +
  
      scale_colour_manual(
              values = c("Low" = "darkgrey","High"  = "#1D9E75"))+
      theme( axis.title.y = element_text(size = 22, angle = 90, margin = margin(r = 10)),
             axis.title.x = element_text(size = 22, margin = margin(5)),
            axis.text.y = element_text(size = 18),
            axis.text.x = element_text(size = 18),
            legend.position = "none"  )  

vt

ggsave("v_t.png", vt, height = 7, width = 7)
 
###### Reviewer Response: Integrated infection prevalence vs. infection prevalence

#See Hite et al. (2016) Ecology.
#See Civitello et al. (2015) Journal of Animal Ecology
 
#Integrated infection prev. vs. **MEAN* annual infeciton prevalence
lake_summary1 %>%
    ggplot(aes(y = mean_inf_prev, x = int_inf_prev, colour = hi_low_fib)) +
        geom_point(size = 4) +
        stat_cor(aes(group =1 ))+
        stat_smooth(method = "glm", aes(group =1), colour = "black") +
        scale_colour_manual(
              values = c("Low" = "darkgrey","High"  = "#1D9E75"))+
        labs(x = "Integrated infection prevalence
(units of proportion × months)", 
             y = "Mean annual infection prevalence
(lake-level, proportion infected)") 


#Integrated infection prev. vs. **MAX** annual infection prevalence
lake_summary1 %>%
    ggplot(aes(y = max_inf_prev, x = int_inf_prev, colour = hi_low_fib)) +
        geom_point(size = 4) +
        stat_cor(aes(group =1 ))+
        stat_smooth(method = "glm", aes(group =1), colour = "black") +
        scale_colour_manual(
              values = c("Low" = "darkgrey","High"  = "#1D9E75"))+
        labs(x = "Integrated infection prevalence
(units of proportion × months)", 
             y = "Maximum annual infection prevalence
(lake-level, proportion infected)") 


        
###
cor_lab <- lake_summary1%>%
  summarise(
    rho = cor(mean_inf_prev, int_inf_prev, method = "pearson", use = "complete.obs"), 
    #pearson, spearman
    p = cor.test(mean_inf_prev, int_inf_prev,  method = "pearson")$p.value) 

cor_lab

#"This index of epidemic size is highly correlated with another index, maximum infection prevalence (R = 0·914, P < 0·00151), which ranged from 18 to 42% in these lakes in from July 2023 to July 2024".
######## 

```

################################### 
# Models
###################################

# PARASITE BIOMASS

Here, I originally attempted to do infection only-- but none of the stats work out because the differences between groups is just so large. So, let's embrace the nature of the data and models both the zeros and non-zeros. 

Parasite biomass cannot be captured by gamms or glmms. 

**E[y] = (1 - zi) × mu**
Where zi is the probability of being a structural zero (uninfected) and mu is the expected burden given the fish is infected. So:

**.epred = (1 - zi) × mu** -- the population-level prediction, averaging over infected and uninfected fish
mu = the conditional prediction -- expected burden given the fish is infected

```{r Bayesian, simplified}

cohort_data_mod <- cohort_data_full %>% #includes babies (but not their fibrosis)
                  filter(!Cohort == "F2") %>%
                  mutate(Cohort = droplevels(Cohort)) 

####################################
######### COMPLEX SPLINES WITH A TP CURVE ##############
####################################

bf_zib_final <- bf(wormbiomass_fishmass ~ hi_low_fib + Cohort +
                        s(Month_num, by = hi_low_fib, bs = "tp", k = 5),
                      zi ~ hi_low_fib + (1 | Lake) +
                        s(Month_num, by = hi_low_fib, bs = "tp", k = 4),
                      phi ~ hi_low_fib + (1 | Lake) +
                        s(Month_num, by = hi_low_fib, bs = "tp", k = 4))  # bumped from 3
                    

pri_zib_final <- c(
  # mu
  prior(student_t(3, 0, 2.5), class = "Intercept"),
  prior(normal(0, 0.6), class = "b"),
  prior(exponential(1), class = "sds"),
  # no sd for mu -- Lake removed from mu
  
  # zi
  prior(student_t(3, 0, 2.5), class = "Intercept", dpar = "zi"),
  prior(normal(0, 0.6), class = "b", dpar = "zi"),
  prior(exponential(1), class = "sds", dpar = "zi"),
  prior(student_t(3, 0, 2.5), class = "sd", dpar = "zi"),
  
  # phi
  prior(normal(log(40), 0.6), class = "Intercept", dpar = "phi"),
  prior(normal(0, 0.3), class = "b", dpar = "phi"),
  prior(exponential(1), class = "sds", dpar = "phi"),
  prior(student_t(3, 0, 2.5), class = "sd", dpar = "phi"))


###### Let's run it
fit_zib_final <- brms::brm(
                bf_zib_final,
                data = cohort_data_mod,
                family = brms::zero_inflated_beta(),
                prior = pri_zib_final,
                chains = 4, iter = 3000, warmup = 1000,
                threads = threading(4),
                control = list(adapt_delta = 0.99, max_treedepth = 14),
                seed = 2025)


################
#### Plotting outcomes ###
################

#This plots EVERYTHING (including zeros!), note scale difference
pred_grid <- expand.grid(
  Month_num = seq(1, 13, by = 0.25),
  hi_low_fib = c("High", "Low"),
  Cohort = "F0")

pred_grid %>%
  add_epred_draws(fit_zib_final, re_formula = NA) %>%
  group_by(Month_num, hi_low_fib) %>%
  summarise(mean = mean(.epred),
            lower = quantile(.epred, 0.025),
            upper = quantile(.epred, 0.975)) %>%
  ggplot(aes(x = Month_num, y = mean,
             color = hi_low_fib, fill = hi_low_fib)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.3, color = NA) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = c("High" = "#2d8f6f", "Low" = "grey50")) +
  scale_fill_manual(values = c("High" = "#2d8f6f", "Low" = "grey50"))

############################################
## Now let's just look at infected individuals
# given this fish is infected, what burden do we expect?

# get mu draws (burden given infected)
mu_draws <- pred_grid %>%
  add_epred_draws(fit_zib_final, re_formula = NA, dpar = "mu")


BAYE<- mu_draws %>%
  group_by(Month_num, hi_low_fib) %>%
  summarise(mean = mean(mu),
            lower = quantile(mu, 0.025),
            upper = quantile(mu, 0.975)) %>%
  ggplot(aes(x = Month_num, y = mean,
             color = hi_low_fib, fill = hi_low_fib)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.1, color = NA) +
  geom_line(linewidth = 1) +
  geom_point(aes(y = wormbiomass_fishmass,
                 x = Month_num),  alpha = 0.05, size = 3,
             data = cohort_data_mod%>% 
             filter(wormbiomass_fishmass > 0)) +
  labs(y = "Parasite burgen (g worm/g host)", x = "")+
  theme( legend.position = "none",
         axis.text.x = element_text(size = 18),
         axis.text.y = element_text(size = 18),
         axis.title.y = element_text(size = 22, 
                          angle =90, margin = margin(r = 15))) +
  scale_x_continuous(breaks = seq(1, 13, by =1))+
  ylim(0, 0.3)+
  scale_color_manual(values = c("High" = "#1D9E75", "Low" = "grey50")) +
  scale_fill_manual(values = c("High" = "#1D9E75", "Low" = "grey50")) 


ggsave("baye.png", BAYE, height = 3, width = 4)


#################################
## Pulling quantities: 
#how do these peaks look between hi_low_fib in INFECTED INFIVIDUALS?
#######

peak_draws <- mu_draws %>%
              group_by(hi_low_fib, .draw) %>%
              slice_max(mu, n = 1) %>%
              dplyr::select(hi_low_fib, .draw, Month_num, mu)


# summarise posterior of peak timing
### HI FIB = 5 (NOVEMBER); LOW-FIB 6.75 (LATE-DEC/JAN)
peak_draws %>%
  group_by(hi_low_fib) %>%
  summarise(
    peak_median = median(Month_num), 
    peak_lower = quantile(Month_num, 0.025),
    peak_upper = quantile(Month_num, 0.975))


###################

########################
# MODEL QUALITY CHECKS
################

#### OUTPUT ####

summary(fit_zib_final) #fit_zib_point, fit_simple
bayesplot::mcmc_intervals(fit_zib_final, regex_pars = "b_") +theme_classic()
bayesplot::mcmc_intervals(fit_zib_final, regex_pars = "^sds_") +theme_classic()

#Let's check the random effect
brms::VarCorr(fit_zib_final)
ranef(fit_zib_final)

#################
#### QUALITY CHECKS 
#################
plot(fit_zib_final) #fit_zib_point, 
neff_ratio(fit_zib_final) #→ all > 0.1 (preferably > 0.4) 
mcmc_neff(neff_ratio(fit_zib_final))


y    <- fit_zib_final$data$wormbiomass_fishmass
yrep <- posterior_predict(fit_zib_final, ndraws = 500) 

# correct PPC for a scalar statistic
ppc_stat(y, yrep, stat = function(z) mean(z == 0)) +
  labs(title = "P(Y = 0) overall")

pp_check(fit_zib_final, type = "stat", stat = "mean")

# LOO cross validation
loo_zib_final <- loo(fit_zib_final)
print(loo_zib_final)

#####################
#DHARMA QUALITY CHECK!
#####################

dharma_zib <- DHARMa::createDHARMa(
  simulatedResponse = t(posterior_predict(fit_zib_final)),
  observedResponse = cohort_data_mod$wormbiomass_fishmass[!is.na(cohort_data_mod$wormbiomass_fishmass)],
  fittedPredictedResponse = apply(posterior_predict(fit_zib_final), 2, median),
  integerResponse = FALSE)

plot(dharma_zib)

#####################
# TEMPORAL AUTOCORRELATION
#####################
library(spdep)

resid <- residuals(model.check)  #residuals from this is from the dharma quality check
#Month infomation
month <- mf$Month_num
mf <- model.frame(fit_zib_final) 

# Aggregate residuals by month
df_resid <- data.frame(month = month, resid = resid)
resid_month <- aggregate(resid ~ month, data = df_resid, FUN = mean)

# Build 2D coordinates for 1D time
coords <- cbind(resid_month$month, rep(0, nrow(resid_month)))

# Nearest-neighbor weights
nb <- knn2nb(knearneigh(coords, k = 1))
lw <- nb2listw(nb, style = "W")

# Moran's I on temporal structure
moran.test(resid_month$resid, lw)
```


#FIBROSIS

Using gamms to look at immune expression across time
```{r Fibrosis timing}
# Data for models
######################

fish_sizes_mod <- fish_sizes %>%
              filter(!Cohort == "F2") %>%
              mutate(fib_int = as.integer(Fibrosis) + 1, 
                     inf_yn = as.integer(ifelse(Total_Worms>0, 1, 0))) %>% #for ocat model
              group_by(Lake, Month.adj) %>%
              mutate(inf_prev_lake_month = sum(inf_yn, na.rm=TRUE)/n()) %>%
              ungroup() %>%
              group_by(Lake, Month.adj, Cohort) %>%
              mutate(inf_prev_lake_month_cohort = sum(inf_yn, na.rm=TRUE)/n()) %>%
              ungroup()

cohort_data_mod <- cohort_data_full %>% #includes babies (but not their fibrosis)
                  filter(!Cohort == "F2") %>%
                 mutate(Cohort = droplevels(Cohort))  %>%  # <- THIS is the key line
                  mutate(fib_int = as.integer(Fibrosis) + 1) #for ocat model

fish_sizes_mod$Cohort <- fct_relevel(fish_sizes_mod$Cohort, "F0")
#############################################
m_fib <- mgcv::gam(
               fib_int ~ Cohort + max_worm_wt+ hi_low_fib *inf_yn+
                          s(Month_num, by = hi_low_fib, 
                            bs = "tp", k = 4) +
                          s(Lake, bs = "re"),
                          family = ocat(R = 5),
                          data   = fish_sizes_mod,
                          knots  = list(Month_num = c(1, 13)),
                          method = "REML")

#max.worm wt n.s., so dropped.
m_fib1 <- mgcv::gam(
                    fib_int ~ Cohort +  hi_low_fib *inf_yn+  
                         # deviation smooths by fibrosis regime
                          s(Month_num, by = hi_low_fib, 
                            bs = "tp", k = 4) + #3
                          s(Lake, bs = "re"),
                          family = ocat(R = 5),
                          data   = fish_sizes_mod,
                          knots  = list(Month_num = c(1, 13)),
                          method = "REML")

#testing the interaction and dropping that prevalence 
AIC(m_fib,  m_fib1) 

#######################################

summary(m_fib1)
plot(m_fib1, pages = 1, scale = 0)
simulationOutput <- simulateResiduals(fittedModel = m_fib1, plot = T)
DHARMa::testDispersion(m_fib1)
DHARMa::testZeroInflation(m_fib1)

############################

# When does fibrosis peak
nd <- expand.grid(
  Month_num  = seq(1, 13, by = 0.1),
  hi_low_fib = levels(fish_sizes_mod$hi_low_fib),
  Cohort     = "F0",  # hold constant
  inf_yn     = mean(fish_sizes_mod$inf_yn, na.rm = TRUE),
  Lake       = factor(levels(fish_sizes_mod$Lake)[1], 
                      levels = levels(fish_sizes_mod$Lake))
) %>%
  mutate(
    hi_low_fib = factor(hi_low_fib, levels = levels(fish_sizes_mod$hi_low_fib)),
    Cohort     = factor(Cohort, levels = levels(fish_sizes_mod$Cohort))
  )

# predicted category probabilities
P <- predict(m_fib1, newdata = nd, type = "response", exclude = "s(Lake)")

# expected fibrosis score -- check your fib_int coding first
k <- 1:5
nd$E_fibrosis <- as.vector(P %*% k)

# peak month per fibrosis group
nd %>%
  group_by(hi_low_fib) %>%
  slice_max(E_fibrosis, n = 1, with_ties = FALSE) %>%
  dplyr::select(hi_low_fib, peak_month = Month_num, peak_E = E_fibrosis)


ggplot(nd, aes(x = Month_num, y = E_fibrosis, color = hi_low_fib)) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = c("High" = "#1D9E75", "Low" = "grey50")) +
  theme_classic()

############################

#Let's generate a table best model is the one that includes individual-level infection!
tab_model(
  m_fib1,
  show.obs = F,
  show.r2 = F,
  show.ci = FALSE,
  show.se = TRUE,
  show.stat = TRUE,
  show.icc = FALSE,
  transform = NULL,
  dv.labels = "Fibrosis score (ordinal GAM)",
  pred.labels = c(
    "Intercept",
    "Cohort [F1]",
    "Immune population [Low-fib]",
    "Indiv. infection (0/1)",
    "Indiv. infection (0/1) × Immune pop. [Low-fib]",
    # threshold parameters and smooths will auto-label
    "s(Month) [High-fib]",
    "s(Month) [Low-fib]",
    "Lake (RE)" ))

```


#TIMING OF TRANSMISSIBLE PARASITES

Here, I want to demonstrate that transmissibility of worms (>0.05g) varies between lakes, time, and stages.Let's take the maximum worm weight in each fish and see if it is of transmissible stage.

```{r MODELS: Transmissible parasites}

#Do low-fibrosis lakes have more transmissible parasites than high-fibrosis lakes?
trans_cor<- cohort_data_full %>%
        mutate(trans_yn = ifelse(max_worm_wt >= 0.05, 1, 0)) %>%
        select(Lake, Fibrosis, Total_Worms, max_worm_wt, hi_low_fib, trans_yn)

kruskal.test(trans_yn ~ hi_low_fib, data = trans_cor)

#Fisher's Exact Test
table_df <- table(trans_cor$hi_low_fib, trans_cor$trans_yn)
table_df

fisher.test(table_df)

######################### 
# What about when do these transmissible parasites occur across time?
######################### 

trans_sum_lake <- trans %>%
                  filter(!Cohort == "F2") %>%
                  group_by(hi_low_fib, Lake, Month.adj, Season, Cohort, Month_num) %>%
                  summarise(tot_trans_stages = sum(trans_worm_yn, na.rm = TRUE),
                            .groups = "drop")

nrow(trans_sum_lake)

library(mgcv)
m1_gamm <- gam(tot_trans_stages ~ hi_low_fib + Cohort +
                 s(Month_num, by = hi_low_fib, bs = "tp", k = 4) +
                 s(Lake, bs = "re"),
               family = nb(),
               data = trans_sum_lake,
               method = "REML")

summary(m1_gamm)
simulationOutput <- simulateResiduals(fittedModel = m1_gamm, plot = T)
DHARMa::testDispersion(m1_gamm)
DHARMa::testZeroInflation(m1_gamm)

plot(m1_gamm, pages = 1, scale = 0)


```


#TRANSMISSION TO BIRDS
```{r CORRELATION: TRANMISSION TO BIRDS}

cor_lab <- infbird %>%
  group_by(hi_low_fib) %>%
  summarise(
    rho = cor(n_species, tot_trans_stages, method = "spearman", use = "complete.obs"), #pearson
    p = cor.test(n_species, tot_trans_stages, method = "spearman")$p.value) 

cor_lab

```


